---
title: "Tutorial 5"
subtitle: "Nonlinearity: interactions"
author: "Andrea Recine"
date: today
lang: en
---

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(modelsummary)
library(fixest)
library(wooldridge)
data("wage1", package = "wooldridge")
```





## Introduction

In this tutorial, we study how to interpret interaction coefficients between two dummy variables, between a dummy variable and a continuous variable, and between two continuous variables. Interactions allow us to model cases in which the effect of one variable depends on the level of another variable.



We return to a previous exercise in which we interpreted coefficients on dummy variables with multiple categories. We created three dummy variables (married women, married men, and single women) and interpreted the coefficients relative to single men, our omitted category:

```{r}
wage1 <- wage1 %>%
  mutate(
    marrmale = ifelse(female == 0 & married == 1, 1, 0),
    marrfemale = ifelse(female == 1 & married == 1, 1, 0),
    singfem = ifelse(female == 1 & married == 0, 1, 0)
  )
```

For simplicity, and to compare this specification with the regression that includes an interaction between $female$ and $married$, we estimate a regression model with these three dummy variables without controlling for other factors such as years of education or work experience.

```{r}
reg_wage_sm1 <- feols(wage ~ marrmale + marrfemale + singfem, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage_sm1), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

How should the coefficients be interpreted?

1. The intercept represents the predicted average hourly wage, in dollars, for the omitted category (**single men**): the expected value of $wage$ when $marrmale$, $marrfemale$, and $singfem$ are all equal to zero.
2. $marrmale$: on average, married men earn 2.815 dollars more per hour than single men.
3. $marrfemale$: on average, married women earn 0.60 dollars less per hour than single men.
4. $singfem$: on average, single women earn 0.55 dollars less per hour than single men.


**Note:** we are not controlling for years of education, years of experience, or other factors.


## Interaction Between Two Dummy Variables

We now create the interaction between two dummy variables: $female$ and $married$. This allows us to examine whether the "return" to marriage depends on gender.


::: {.callout-note title="Specifying interactions with `fixest`"}

In `fixest`, interactions can be specified more compactly with the `*` operator, which automatically includes the main effects and the interaction term. Therefore:

```{r}
reg_wage_marrfe <- feols(wage ~ female * married, data = wage1, vcov = "hetero")
```

If we wanted to specify only the interaction term without the main effects, we could use the `:` operator as follows:

```{r}
reg_wage_marrfe <- feols(wage ~ female + married + female:married, data = wage1, vcov = "hetero")
```

These two commands produce the same regression model.

:::

```{r}
reg_wage_marrfe <- feols(wage ~ female * married, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage_marrfe), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```



The regression results show that the interaction between gender and marital status is statistically significant (compute the critical value).


The coefficient on $female$ represents the wage difference between unmarried women and unmarried men. On average, unmarried women earn 0.556 dollars less per hour than unmarried men.


The coefficient on $married$ represents the effect of marriage on hourly wage for **men**. On average, married men earn 2.815 dollars more per hour than unmarried (single) men.


The coefficient on the interaction $female \times married$ represents the difference in the marriage effect between women and men. The effect of marriage on hourly wage is 2.861 dollars lower for women than for men.


On average, married women earn 0.046 dollars less per hour than unmarried women ($2.815 - 2.861 = -0.046$).


::: {.callout-note title="Note"}
**We can use the estimated regression to compute the average hourly wage for each of the four possible combinations:**

- The intercept, equal to 5.168, represents the average hourly wage for single (unmarried) men, that is, when $female$ = 0 and $married$ = 0.

- When $female$ = 0 and $married$ = 1, we obtain 7.983 (5.168 + 2.815), which represents the average hourly wage in dollars for married men.

- When $female$ = 1 and $married$ = 0, we obtain 4.612 (5.168 - 0.556), which represents the average hourly wage for single (unmarried) women.

- When $female$ = 1 and $married$ = 1, we obtain 4.566 (5.168 - 0.556 + 2.815 - 2.861), which is the average hourly wage for married women.
:::

**Note:** we are not controlling for other factors such as years of education.




### Interaction Between $male$ and $married$

We create a new dummy variable, $male$, equal to 1 for men and 0 for women.

```{r}
wage1 <- wage1 %>%
  mutate(male = 1 - female)
```

```{r}
reg_wage_marrmal <- feols(wage ~ male * married, data = wage1, vcov = "hetero")
modelsummary(list("Wage (Female = 1)" = reg_wage_marrfe, "Wage (Male = 1)" = reg_wage_marrmal), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


In this case (the second regression column), the coefficient on $married$ represents the effect of marriage on hourly wage for women. On average, married women earn 0.046 dollars less per hour than unmarried women.


The coefficient on $male$ represents the wage difference between unmarried men and unmarried women. On average, unmarried men earn 0.556 dollars more per hour than unmarried women.


The intercept represents the average wage for single (unmarried) women, that is, when $male$ = 0 and $married$ = 0.


The coefficient on the interaction $male \times married$ represents the difference in the marriage effect between men and women. The effect of marriage on hourly wage is 2.861 dollars higher for men than for women.


## Interaction Between a Dummy Variable and a Continuous Variable

In this tutorial, we consider the case that the textbook (Stock and Watson, p. 220) describes as different slopes and intercepts. We therefore include $educ$, $female$, and their interaction.


```{r}
# First model: regression of wage on educ, female, and their interaction
reg_wage_educfe <- feols(wage ~ educ * female, data = wage1, vcov = "hetero")


# Results
modelsummary(list("Wage" = reg_wage_educfe), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


- The coefficient on $educ$ represents the effect of an additional year of education on wage for men (when $female$ = 0). The effect of an additional year of education for men is 0.539 dollars per hour.

- The coefficient on $female$ represents the wage difference between women and men with zero years of education ($educ$ = 0).

- The coefficient on $educ \times female$ represents the difference in the effect of education between women and men. Women receive an education premium that is about 0.086 dollars per hour lower than the one received by men.

- The effect of an additional year of education for women is equal to $\beta_1 + \beta_3$, that is, $0.539 - 0.086 = 0.453$ dollars.


Indeed, $0.453 - 0.539 = -0.086$ is exactly the interaction coefficient.

```{r}
# install.packages("marginaleffects")
library(marginaleffects)
plot_predictions(reg_wage_educfe,
                condition = c("educ", "female")) +
  labs(title = "Predicted Wage by Education and Gender",
       x = "Years of education",
       y = "Predicted wage")
```


The plot shows that predicted wage increases with years of education for both men and women, but the slope is steeper for men than for women (although the difference is not very large).



## Interaction Between Two Continuous Variables

We consider the interaction between $educ$ and $exper$ to determine whether the effect of an additional year of education on wage depends on years of work experience (or vice versa).

```{r}
reg_wage_educexper <- feols(wage ~ educ * exper, data = wage1, vcov = "hetero")

reg_wage_educexper2 <- feols(wage ~ educ * exper + female + tenure, data = wage1, vcov = "hetero")
# Results
modelsummary(list("Wage" = reg_wage_educexper, "Wage" = reg_wage_educexper2), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


- The coefficient on $educ$ represents the marginal effect of an additional year of education on wage when experience ($exper$) is zero.

- The coefficient on $exper$ represents the marginal effect of an additional year of experience on wage when education ($educ$) is zero.


The interaction coefficient is positive: this means that the effect of education increases with experience (or vice versa).


What is the marginal effect of an additional year of education at different levels of experience?


**Marginal effect of an additional year of education:**

$$
\frac{\partial wage}{\partial educ} = \beta_1 + \beta_3 \cdot exper
$$


To estimate the partial effect, we consider relevant values of $exper$, such as the median, the mean, or the third quartile.

```{r}
datasummary(exper ~ Min + P25 + Median + Mean + P75 + Max, data = wage1)
```

Using the median: $0.602 + 0.002 \times 13.5 = 0.629$

- For an individual with 13.5 years of experience (the median), the marginal effect (association) of an additional year of education is about 0.629 dollars.


- For an individual with 26 years of experience (the third quartile), an additional year of education is associated with an average increase in hourly wage of about $0.602 + 0.002 \times 26 = 0.65$ dollars.


## Deviation from the Mean

### Interaction Between Continuous Variables: Centering


Define:


```{r}
wage1 <- wage1 %>%
  mutate(
    educ_center = educ - mean(educ, na.rm = TRUE),
    exper_center = exper - mean(exper, na.rm = TRUE)
  )
```



```{r}
reg_wage_educexper_center <- feols(wage ~ educ_center * exper_center, data = wage1, vcov = "hetero")
# Results
modelsummary(list("Wage" = reg_wage_educexper_center), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


Now the coefficient associated with $educ\_center$ represents the effect of an additional year of education on wage when experience ($exper$) is equal to its mean.

An additional year of education is associated with an increase in hourly wage of about 0.637 dollars for an individual with mean experience (17.02 years).


### Interaction Between a Dummy Variable and a Continuous Variable: Centering the Continuous Variable

Centering the continuous variable can also be useful when interacting a dummy variable with a continuous variable:

```{r}
# Model with centered educ and interaction with female (reg_wage_educfe already defined above)
reg_wage_educfe_center <- feols(wage ~ educ_center * female, data = wage1, vcov = "hetero")

# Results
modelsummary(list("Wage" = reg_wage_educfe, "Wage" = reg_wage_educfe_center), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


**Without centering:**

- The coefficient on $educ$ represents the effect of an additional year of education on wage for men (when $female$ = 0). The effect of an additional year of education for men is 0.539 dollars per hour.

- The coefficient on $female$ represents the wage difference between women and men when education is zero ($educ$ = 0).

**Second model (with centering):**

- The coefficient on $female$ represents the wage difference between women and men when years of education ($educ$) are at their mean value.
