---
title: "Tutorial 2"
subtitle: "Multiple Regression"
author: "Andrea Recine"
date: today
lang: en
description: "Multiple regression: partialling out, dummy variables, and joint hypothesis tests (wage1)."
code-links:
  - text: R Script
    icon: file-code
    href: tutorial2_script.R
---

## Multiple Regression

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(tinytable)
library(modelsummary)
```


```{r}
#| warning: false
library(wooldridge)
```

```{r}
data("wage1", package = "wooldridge")
```


Multiple regression explicitly allows us to control for other variables that affect the dependent variable and may be correlated with the regressor.

**Idea**: By including the relevant variables in the model, and therefore removing them from the error term, we make it more plausible that the remaining factors in the error term are not correlated with the model variables.

In the previous [Tutorial](../tutorial1/tutorial1.qmd), we estimated the following simple regression model:

$$
wage_i = \beta_0 + \beta_1 educ_i + u_i
$$

where $wage$ is measured in dollars per hour (in 1976 dollars) and $educ$ is years of education.


We now estimate two additional models:

1. we add $exper$ which represents years of experience in the labor market

2. in addition to $educ$ and $exper$, we add $tenure$, which represents the number of years with the current employer.

Therefore:

$$
wage_i = \beta_0 + \beta_1 educ_i + \beta_2 exper_i + u_i
$$
$$
wage_i = \beta_0 + \beta_1 educ_i + \beta_2 exper_i + \beta_3 tenure_i + u_i
$$


```{r}
# Regression
library(fixest)
reg_wage1 <- feols(wage ~ educ, data = wage1, vcov = "hetero")
reg_wage2 <- feols(wage ~ educ + exper, data = wage1, vcov = "hetero")
reg_wage3 <- feols(wage ~ educ + exper + tenure, data = wage1, vcov = "hetero")
```

```{r}
#| label: tab1
#| tbl-cap: "OLS estimation results. Dependent variable $:$ wage"
modelsummary(list("Wage" = reg_wage1, "Wage" = reg_wage2, "Wage" = reg_wage3), gof_omit = "AIC|BIC|RMSE")
```

**Interpretation of the coefficients**: When interpreting coefficients, we consider the effect of one variable holding all other variables constant.


We are still interested in the effect of $educ$ on $wage$, holding the other factors fixed.


**Second column**: An additional year of education is associated, on average, with a 0.64 dollar increase in hourly wage, holding years of labor-market experience constant. Individuals with more years of education tend to enter the labor market later and therefore, holding age constant, have less work experience. This explains why $educ$ and $exper$ are negatively correlated.



::: {.callout-note collapse="false"}
## Correlation Between $educ$ and $exper$

```{r}
cor(wage1$educ, wage1$exper)
```


The omitted-variable bias in the simple model `reg_wage1` depends on two factors: the effect of $exper$ on $wage$ and the correlation between $exper$ and $educ$. If both differ from zero, the bias differs from zero.

The effect of $exper$ on $wage$ is positive: more experience is associated with higher wages. The covariance, and therefore the correlation, between $educ$ and $exper$ is negative ($Cov(educ, exper) < 0$): individuals with more years of education tend to have less experience.

The bias in the coefficient on $educ$ is therefore **negative**: the simple model **underestimates** the effect of education on wage.


:::


Alternatively, we may be interested in the effect of experience on hourly wage, holding years of education constant. An additional year of labor-market experience ($exper$) is associated, on average, with a 0.070 dollar increase in hourly wage, holding years of education constant.


In this case, the key assumption (zero conditional mean) is $E(u_i | educ_i, exper_i) = 0$.

---

**Third model**: An additional year of education is associated, on average, with a 0.59 dollar increase in hourly wage, holding years of labor-market experience and years with the same employer ($tenure$) constant.


The key assumption (zero conditional mean) is $E(u_i | educ_i, exper_i, tenure_i) = 0$.


::: {.callout-note collapse="true"}
## Partialling Out (Frisch-Waugh-Lovell Theorem)

How do we obtain the coefficient on `educ` in a multiple regression? The Frisch-Waugh-Lovell theorem shows that it is equivalent to the following procedure:

1. Regress `wage` on the other variables (`exper`, `tenure`)
   and compute the residuals: the part of wage *not* explained
   by experience and tenure.
2. Regress `educ` on the same variables and compute the residuals:
   the part of education *not* explained by experience and tenure.
3. Regress the residuals from step 1 on the residuals from step 2:
   the coefficient obtained is exactly $\hat{\beta}_{educ}$
   from the multiple regression.

In other words, the coefficient on `educ` captures the association
between wage and education after removing the effects of the
other variables, hence the term "partialling out."
:::


### Including a dummy variable:

```{r}
# Regression
reg_wage4 <- feols(wage ~ educ + exper + tenure + female, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage3, "Wage" = reg_wage4), gof_omit = "AIC|BIC|RMSE")
```

In this regression, we are also controlling for gender (dummy variable $female$).


If $E(u_i | educ_i, exper_i, tenure_i, female_i) = 0$, then $\beta_1$ represents the causal effect of an additional year of education on hourly wage, **holding experience, years with the same employer, and gender constant**.


---

### Changing units of measurement:

Suppose wage is expressed on a monthly basis for full-time workers who work 7 hours per day, 5 days per week, and 4 weeks per month. This gives $7 \times 5 \times 4 = 140$ hours per month.

We create a new variable wage_monthly = $wage$ $\times$ 140

```{r}
wage1 <- wage1 %>%
  mutate(wage_monthly = wage*140)


reg_wage_dummy3 <- feols(wage_monthly ~ educ + exper + tenure + female, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage4, "Wage Monthly" = reg_wage_dummy3), gof_omit = "AIC|BIC|RMSE")
```


**Hourly wage**: An additional year of education is associated, on average, with a 0.572 dollar increase in hourly wage, holding gender, experience, and $tenure$ constant.


**Monthly wage**: An additional year of education is associated, on average, with an 80 dollar increase in monthly wage, holding gender, experience, and $tenure$ constant.


What if $educ$ were measured in months of education?

In [Tutorial 1](../tutorial1/tutorial1.qmd), we saw that expressing $educ$ in months by multiplying it by 12 leaves the intercept unchanged and divides the coefficient on $educ$ by 12.


The coefficient on education measured in months will equal 0.048 dollars ($\frac{0.572}{12}$). An additional month of education is associated, on average, with a 0.048 dollar increase in hourly wage, holding gender, experience, and $tenure$ constant. The remaining coefficients, including the intercept and the dummy-variable coefficient, remain unchanged.



---

## Dummy Variable: Wage and Gender:

In [Tutorial 1](../tutorial1/tutorial1.qmd), we estimated the following regression:

$$
wage_i = \beta_0 + \beta_1 female_i + u_i,
$$

```{r}
reg_wage_dummy <- feols(wage ~ female, data = wage1, vcov = "hetero")
reg_wage_dummy2 <- feols(wage ~ female + educ + exper + tenure, data = wage1, vcov = "hetero")
modelsummary(list("Wage (Tutorial 1)" = reg_wage_dummy, "Wage" = reg_wage_dummy2), gof_omit = "AIC|BIC|RMSE")
```

The results of the first regression show that the average wage of men (represented by the intercept) is about 7.1 dollars (in 1976 dollars) per hour.

The coefficient on the female dummy variable is -2.51 dollars. On average, women earn about 2.51 dollars less per hour than men. The average hourly wage for women is about 4.58 dollars ($7.1 - 2.51$).


In the second regression, we control for differences in education, experience, and $tenure$.


In this second regression, the negative intercept for men has little practical meaning because no individual in the sample has zero years of $educ$, $exper$, and $tenure$. The coefficient on $female$, however, represents the average difference in hourly wage between a woman and a man, holding $educ$, $exper$, and $tenure$ constant.


Because we control for $educ$, $exper$, and $tenure$, the 1.81 dollar hourly-wage gap cannot be explained by differences in the average levels of education, experience, and tenure between men and women. The 1.81 dollar difference is attributable to gender or to factors associated with gender that are not controlled for in the regression.

---

## Dummy Variables with Multiple Categories


We estimate a model that allows for wage differences across four groups: married men, married women, single men, and single women.

To interpret the dummy-variable coefficients, we must choose a reference group.

We then define dummy variables for the other three groups.

**Dummy variables**: We use single men as the omitted category and create the following variables:

- $marrmale$ equals 1 for married men.

- $marrfemale$ equals 1 for married women.

- $singfem$ equals 1 for single (unmarried) women.

- $singmale$ equals 1 for single (unmarried) men. This is our reference group. We create the dummy variable for single men but do not include it in the regression.



```{r}
wage1 <- wage1 %>%
  mutate(
    marrmale = ifelse(female == 0 & married == 1, 1, 0),
    marrfemale = ifelse(female == 1 & married == 1, 1, 0),
    singfem = ifelse(female == 1 & married == 0, 1, 0),
    singmale = ifelse(female == 0 & married == 0, 1, 0)
  )
```


What would happen if we also included $female$? We would obtain perfect collinearity, known as the dummy-variable trap.

When several dummy variables must be created, for example for regions, the `fastDummies` package can be useful.


::: {.callout-note title="Recall"}
The mean of a dummy variable is the proportion of observations equal to 1 in the sample:
:::

```{r}
statdummy <- wage1 %>%
  summarise(
    marrmale_mean = mean(marrmale, na.rm = TRUE),
    marrfemale_mean = mean(marrfemale, na.rm = TRUE),
    singfem_mean = mean(singfem, na.rm = TRUE),
    singmale_mean = mean(singmale, na.rm = TRUE)
  )

tt(statdummy, width = 1, digits = 3, caption = "Means of the dummy variables")
```

The sum $0.357 + 0.251 + 0.228 + 0.163$ equals 1, as expected, because the four dummy variables represent mutually exclusive and collectively exhaustive groups.


```{r}
reg_wage_sm <- feols(wage ~ marrmale + marrfemale + singfem + educ + exper + tenure, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage_sm), gof_omit = "AIC|BIC|RMSE")
```

**Interpretation of the coefficients:**

- On average, a married man earns 1.822 dollars more per hour than a single man, holding years of education, years of experience, and years with the current employer ($tenure$) constant.

- On average, a married woman earns 0.882 dollars less per hour than a single man, holding years of education, years of experience, and $tenure$ constant.

- On average, a single woman earns 0.309 dollars less per hour than a single man, holding years of education, years of experience, and $tenure$ constant.


::: {.callout-caution title="Warning"}
Are these coefficients statistically significant at the 5% level? Also compute their confidence intervals.
:::


---

### The Dummy-Variable Trap

As in Tutorial 1, we create the variable $male$.
```{r}
wage1 <- wage1 %>%
  mutate(male = 1 - female)
```

We regress $wage$ on $marrmale$, $marrfemale$, $singfem$, $female$, and $male$.

```{r}
reg_wage_sm_dummy <- feols(wage ~ marrmale + marrfemale + singfem + female + male, data = wage1, vcov = "hetero")
reg_wage_sm_dummy
```

::: {.callout-important title="Warning"}
$female$ and $male$ are perfectly collinear.

This is because:

- $female$ = $marrfemale$ + $singfem$

- $male$ = $marrmale$ + $singmale$ (where single men are the omitted **reference group**).


The variable $male$ identifies all men and is the sum of married men (captured by $marrmale$) and single men (the reference category, which is not included in the regression).

R automatically removes the perfectly collinear variables.
:::


---


### Dummy Variables for Geographic Regions


The `wage1` dataset divides the United States into four major regions defined by the _US Census Bureau_: **Northeast**, **North Central / Midwest**, **South**, and **West**.


The dataset contains three dummy variables, `northcen`, `south`, and `west`, while the Northeast is the **omitted reference group**.

::: {.callout-note collapse="true"}
## Historical note

The Census Bureau renamed the "North Central" region "Midwest" in 1984. Because the `wage1` dataset is from 1976, it retains the older designation, which explains the variable name `northcen`.
:::


```{r}
wage1 |>
  summarise(
    n_northcen  = sum(northcen),
    n_south     = sum(south),
    n_west      = sum(west),
    n_northeast = sum(1 - northcen - south - west)
  ) |>
  tt(caption = "Observations by region")
```


The sum across these four categories equals the total number of observations, as expected, because the categories are mutually exclusive and exhaustive.

Before estimating the model, we examine the *unadjusted* wage differences across regions:

```{r}
wage1 |>
  mutate(
    region = case_when(
      northcen == 1 ~ "North Central",
      south    == 1 ~ "South",
      west     == 1 ~ "West",
      TRUE          ~ "Northeast"
    )
  ) |>
  group_by(region) |>
  summarise(
    n = n(),
    mean_wage   = mean(wage),
    median_wage = median(wage),
    mean_educ   = mean(educ)
  ) |>
  arrange(desc(mean_wage)) |>
  tt(caption = "Descriptive statistics by region") |>
  format_tt(digits = 3) |>
  theme_tt("striped")

```

These are unadjusted differences. In the multiple regression, we control for education, experience, gender, and tenure to determine whether wage differences across regions remain after holding these characteristics constant.

We estimate the following regression:

```{r}
reg_region <- feols(
  wage ~ educ + exper + tenure + female + northcen + south + west,
  data = wage1, vcov = "hetero"
)

modelsummary(
  list("Wage" = reg_region),
  gof_omit = "AIC|BIC|RMSE|R2 Adj.",
  title = "Regression with Regional Dummies (Northeast = Reference Group)"
)
```



Each coefficient on a regional dummy variable is interpreted **relative to the reference group (Northeast)**:


- $northcen$: a worker in the North Central region (now called the Midwest) earns, on average, 0.621 dollars **less** per hour than a worker in the Northeast, holding years of education, experience, gender, and $tenure$ constant.


- $south$: a worker in the South earns, on average, 0.572 dollars **less** per hour than a worker in the Northeast, holding years of education, experience, gender, and $tenure$ constant.

- $west$: a worker in the West earns, on average, 0.571 dollars **more** per hour than a worker in the Northeast, holding years of education, experience, gender, and $tenure$ constant.



::: {.callout-important title="The Regional Dummy-Variable Trap"}
If we also included a `northeast` dummy variable (constructed as
`1 - northcen - south - west`), we would obtain **perfect collinearity**,
just as in the earlier case with `marrmale + marrfemale + singfem + female + male`.
R would automatically remove one of the four dummy variables.

With $k$ mutually exclusive and exhaustive categories, the sum of the $k$ dummy variables always equals 1 for every observation. This creates perfect collinearity with the intercept, so we include only $k-1$ dummy variables (or all $k$ dummy variables without an intercept).
:::



## Testing Joint Hypotheses

Consider the regression `reg_wage_sm`. We test the null hypothesis that there are no wage differences across the categories after controlling for years of education, experience, and years with the same employer:

That is:
$$
H_0: \beta_1 = \beta_2 = \beta_3 = 0
$$


We can use the `wald()` function from `fixest`:

```{r}
wald(reg_wage_sm,
     keep = "marrmale|marrfemale|singfem")
```

More often, we will use `linearHypothesis()` from the `car` package:

```{r}
# install.packages("car")
library(car)
# Hypothesis test (Wald)
linearHypothesis(reg_wage_sm,
                 c("marrmale = 0", "marrfemale = 0", "singfem = 0"))
```

This is the $\chi^2$ (Wald) statistic based on an asymptotic distribution.


To obtain the $F$ statistic, specify `test = "F"` in `linearHypothesis()`:

```{r}
linearHypothesis(reg_wage_sm,
                 c("marrmale = 0", "marrfemale = 0", "singfem = 0"), test = "F")
```



::: {.callout-note title="Recall"}

| $q$ (restrictions) | Critical value $\chi^2$ ($\alpha = 0.05$) | Critical value $F$ ($\alpha = 0.05$) |
|------------------|-----------------------------|-----------------------------|
| 1                | 3.84                        | 3.84                        |
| 2                | 6.00                        | 3.00                        |
| 3                | 7.80                        | 2.60                        |

*Note:* For large $n$, the $F$ statistic has a $\chi^2/q$ distribution.
:::

In our case:

- 24 > 2.60

- 72.04 > 7.80

We therefore reject the null hypothesis that there are no wage differences across the categories.

There is a simple formula for the $F$ statistic that is valid only under **homoskedasticity**, but it helps illustrate what the statistic does.

Under homoskedasticity, the classical $F$ test proceeds as follows:

- Estimate two regressions: one under the null hypothesis (the restricted regression) and one under the alternative hypothesis (the unrestricted regression).

- Compare the fit of the two regressions using their $R^2$ values. If the unrestricted model fits sufficiently better, reject the null hypothesis.


The classical $F$ statistic is historically important and useful for intuition, but it is not valid in the presence of heteroskedasticity.
