---
title: Tutorial 10
subtitle: Regression with instrumental variables (Part 2)
format: clean-typst
author:
  - name: Econometrics I
    email:
    affiliations: Sapienza University of Rome
date: today
date-format: long
brand:
  typography:
    fonts:
      - family: Roboto
        weight: [light, bold]
        source: google
execute:
  echo: true
  warning: false
  message: false
  error: false
---

# Angrist and Evans (1998)

## Dataset "labsup"


```{r}
library(tidyverse)
library(wooldridge)
library(modelsummary)
data("labsup", package = "wooldridge")
```

- $hours$: mother's weekly hours worked
- $kids$: number of children
- $educ$: years of education
- $age$: age in years
- $black$: dummy
- $hispan$: dummy
- $samesex$: **dummy variable = 1 if the first two children are of the same sex**
- $multi2nd$: **dummy variable = 1 if the second birth produced twins**


## OLS Regression (Exercise)

Regression of (weekly) hours worked on the number of children

```{r}
library(fixest)
ols_labsup <- feols(hours ~ kids + educ + age + black + hispan, data = labsup, vcov = "hetero")
```

- How should the coefficients on $kids$ and $educ$ be interpreted?

## OLS Regression (Exercise)

::: {.slide .center .large}
```{r}
#| echo: false
#| results: 'asis'
modelsummary(list("Hours" = ols_labsup), gof_omit = "AIC|BIC|RMSE|R2 Adj.|R2")
```

:::

- Having one additional child is associated with a reduction, on average, of 2.254 weekly hours worked, holding the other variables constant
- An additional year of education is associated with an increase, on average, of 0.509 hours worked, holding the other variables constant.


## Regression with instrumental variables

### Validity

- The instrumental variable (or "instrument") $Z$ must satisfy the following conditions:

  1. **Relevance:**  $cor(Z_i, X_i) \neq 0$
  2. **Exogeneity:** $cor(Z_i, u_i) = 0$


- In the case of Angrist and Evans (1998), the instrument $samesex$:
  1. Having the first two children of the same sex must be associated with a larger number of children
  2. Having the first two children of the same sex must not directly affect labor supply (the sex composition of the children—two boys or two girls—should not be associated with a different labor supply than, for example, one boy and one girl)




## Regressions with instrumental variables

### Validity

The first condition can be tested (as we will see in the first stage). The second concerns the covariance between $Z$ and the unobserved error $u$. Generally, we cannot test this assumption, and in many cases we assume $Cov(Z, u) = 0$ based on reasoning (for example, theory). We will test the "overidentifying restrictions".

## TSLS

```{r}
iv_labsup <- feols(hours ~ educ + age + black + hispan | kids ~ samesex, data = labsup, vcov = "hetero")
```

- Estimates an IV model in which $kids$ is the endogenous variable, instrumented with $samesex$.

- In the first stage, it regresses the endogenous variable on the instrument ($kids$ on $samesex$) and the other exogenous variables

- In the second stage, it regresses the dependent variable on the predicted values from the first stage ($hours$ on $\hat{kids}$ and the other variables)

- The standard errors account for the first-stage estimation


## TSLS Results ($samesex$ instrument)

::: {.slide .center .large}
```{r}
#| echo: false
#| results: 'asis'
modelsummary(list("Hours" = iv_labsup), gof_omit = "AIC|BIC|RMSE|R2 Adj.|R2")
```

:::


## TSLS (First Stage)
```{r}
summary(iv_labsup , stage = 1)
```

## TSLS (First Stage)

```{r}
fs_labsup <- feols(kids ~ samesex + educ + age + black + hispan, data = labsup, vcov = "hetero")
fs_labsup
```

## TSLS (Second Stage)

```{r}
labsup$kids_hat <- predict(fs_labsup)
feols(hours ~ kids_hat + educ + age + black + hispan, data = labsup, vcov = "hetero")
```

- With this procedure we obtain the same results as `iv_labsup`. But the standard errors are **not** correct (they do not account for the first-stage estimation)


## Instrument Relevance

- We compute the $F$ statistic to test the hypothesis that the coefficients on the instruments are all 0 in the **first stage** of the TSLS regression

- An $F$ statistic $< 10$ indicates that the instruments are weak (Staiger and Stock, 1997; Stock and Yogo, 2005),

- It is not the overall $F$ statistic; rather, we jointly test that the coefficients on the instruments are equal to zero

- If the first-stage Wald statistic is less than $m \times 10$, then the set of instruments is weak. **Note: Wald = $m \times F$**

- Some studies suggest higher critical values or other tests (Montiel Olea and Pflueger, 2013; Kleibergen-Paap rk statistics)


## Instrument Relevance

```{r}
library(car)
linearHypothesis(fs_labsup , "samesex=0")
```

- The Wald statistic equals 48.758 > 1*10


## Instrument Relevance (F)

```{r}
linearHypothesis(fs_labsup , "samesex=0", test="F")
```


## `multi2nd` (twin second birth) as an instrument

```{r}
iv_labsup <- feols(hours ~ educ + age + black + hispan | kids ~ multi2nd, data = labsup, vcov = "hetero")
iv_labsup
```

Relevance and exogeneity in this case:

- **Relevance:** a twin second birth must be associated with a larger number of children
- **Exogeneity:** Having a twin second birth must not directly affect the mother's labor supply



## Instrument Relevance (2 instruments)
We compute the first stage of the regression:

```{r}
fs_labsup_overid <- feols(kids ~ samesex + multi2nd + educ + age + black + hispan, data = labsup, vcov = "hetero")
```

We test the relevance of both instruments: $multi2nd$ and $samesex$


234.72 > 2*10

## Instrument Relevance (2 instruments)

```{r}
linearHypothesis(fs_labsup_overid, c("samesex=0", "multi2nd=0"))
```



## Overidentifying Restrictions
###  Sargan's J

When the number of available instruments $m$ is greater than the number of endogenous variables $k$, the model is overidentified.

- $H_0$: all instruments are exogenous
- $H_1$: at least one instrument is endogenous


## Sargan's J

### Procedure

1. Estimate the TSLS regression
2. Obtain the residuals of the TSLS regression:
   $\hat{u}_i = Y_i - \hat{Y}_i$
3. Regress the residuals $\hat{u}_i$ on the instruments $Z_1, \dots, Z_m$ and the exogenous variables $W_1, \dots, W_r$
4. If the instruments were exogenous, the coefficients on the instruments in the regression of $\hat{u}_i$ should be equal to zero (statistically insignificant)
5. The statistic is computed as $J = mF$. Hence $J = Wald$
6. In large samples it is distributed as a chi-squared with $m-k$ degrees of freedom ($\chi_{m-k}$)

$m-k$ is the degree of overidentification ($k$ is the number of endogenous regressors)


## Sargan's J

```{r}
iv_labsup_overid <- feols(hours ~ educ + age + black + hispan | kids ~ samesex + multi2nd, data = labsup, vcov = "hetero")
```

```{r}
labsup <- labsup |> mutate(uhat = hours - iv_labsup_overid$fitted.values)
Jlm <- feols(uhat~ samesex + multi2nd + educ +age + black + hispan, data = labsup, vcov = "iid")
```


## Sargan's J

```{r}
linearHypothesis(Jlm, c("samesex=0", "multi2nd=0"))
```


## Sargan's J

::: {.callout-important title="Warning"}

The overidentification test returns a statistic
$J=0.583$.

The $J$ statistic is distributed as a $\chi^2_{(m - k)}$. In our case $m -k = 1$​. The value to compare against is **not** 6 but **3.84** ($\alpha = 0.05$)

Thus 0.583 < 3.84. We do not reject $H_0$. The correct p-value is:

```{r}
pchisq(0.583, df = 1, lower.tail = FALSE)
```

or, equivalently, `1 - pchisq(J$Chisq[2], df = 1)`; thus we do not reject the null hypothesis that the instruments are exogenous. **We do not reject $H_0$. All instruments are exogenous.**
:::


## IV Regression with Two Instruments


::: {.slide .center .large}
```{r}
#| echo: false
#| results: 'asis'
modelsummary(list("Hours" = iv_labsup_overid), gof_omit = "AIC|BIC|RMSE|R2 Adj.|R2|Num\\.Obs", stars =TRUE)
```

:::



