---
title: "Tutorial 4"
subtitle: "Nonlinearity: polynomials"
author: "Andrea Recine"
date: today
lang: en

format:
  html:
    # Layout
    theme: flatly
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: right
    number-sections: true

    # Code
    code-fold: false
    code-tools: true
    code-copy: true
    code-block-bg: true
    code-block-border-left: false
    highlight-style: github

    # Output
    embed-resources: true

    # Typography
    fontsize: 1.1em
    linestretch: 1.6

execute:
  warning: false
  message: false
  echo: true

knitr:
  opts_chunk:
    comment: "#>"
    fig.align: "center"
    fig.width: 7
    fig.height: 4.5

---





## Setup:

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(tinytable)
library(modelsummary)
library(fixest)
library(wooldridge)
```



## Introduction


We are interested in the effects of years of experience on hourly wage ($wage$). A quadratic polynomial is often used in economic applications to capture increasing and decreasing marginal effects.



In this case, $\beta_1$ is positive and $\beta_2$ is negative.

```{r}
reg_wage_exper <- feols(wage ~ exper + exper^2, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage_exper), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

::: {.callout-important}
## Syntax `^2` vs `I(^2)`

In R formulas, the `^` operator **does not mean "squared"**: it indicates interactions up to a given order. For this reason, in *almost all* estimation packages (`lm()`, `glm()`, `lme4`, ...) a quadratic term must be written inside `I()` ("inhibit"), which tells R to treat it as an arithmetic operation:

```r
lm(wage ~ exper + I(exper^2), data = wage1)
```

If we wrote `lm(wage ~ exper + exper^2, ...)`, the regression would run without errors, but it would **silently estimate only** $wage$ on $exper$ (without the quadratic term)!

`fixest` (the library we use with `feols()`) is an exception: it automatically recognizes `^` as exponentiation, so `exper^2` is **equivalent** to `I(exper^2)`.
:::


```{r}
# 1. Predicted values
wage1 <- wage1 %>%
  mutate(wageexp_pred = predict(reg_wage_exper))

# 2. Filter to exper <= 40
wage_cut <- wage1 %>% filter(exper <= 40)

# 3. Find the maximum point of the wageexp_pred curve
punto_max <- wage_cut %>%
  filter(wageexp_pred == max(wageexp_pred, na.rm = TRUE))

# 4. Plot
ggplot(wage_cut, aes(x = exper)) +
  geom_line(aes(y = wageexp_pred), color = "#ff7f0e", linewidth = 1.5) +              # estimated curve
  geom_smooth(aes(y = wage), method = "lm", se = FALSE, color = "#1f77b4", linewidth = 1.5) + # OLS line
  geom_point(data = punto_max, aes(y = wageexp_pred), color = "#2ca02c", size = 5) +
  labs(
    title = "Comparison: Linear vs Quadratic Polynomial",
    x = "Years of Experience",
    y = "Hourly wage in $"
  ) +
  theme_minimal()
```


```{r}
reg_wage_exper2 <- feols(wage ~ exper + exper^2 + educ + tenure, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage_exper, "Wage" = reg_wage_exper2), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

### Interpretation of the coefficients:

**Consider the first regression (first column):** we cannot interpret the coefficient associated with $\beta_1$ while holding the coefficient associated with $\beta_2$ constant.


For small changes in $x$, the _ceteris paribus_ effect is described by the marginal effect, so:

We have:

$$
y = \beta_0 + \beta_1 x + \beta_2  x^2 + u
$$


The estimated equation is:

$$
\hat{y} = \hat{\beta_0} + \hat{\beta_1} x + \hat{\beta_2} x^2
$$


We obtain the following approximation in changes:

$$
\Delta \hat{y} \approx  (\hat{\beta_1} + 2  \hat{\beta_2} x) \Delta x
$$

and therefore the marginal effect is:

$$
\Delta \hat{y}/ \Delta x \approx  \hat{\beta_1} + 2  \hat{\beta_2} x
$$

In our example:

- The marginal effect for a worker with one year of experience is: $0.298 - (2* 0.006) * 1 = 0.286$

- The marginal effect at 10 years of experience: $0.298 - 2(0.006) *10 = 0.178$



A discrete change in $\Delta x$ (from 1 to 2 years of experience):

$$
\Delta \hat{Y} = (\hat{\beta_0} + \hat{\beta_1} \cdot 2 + \hat{\beta_2} \cdot 2^{2}) - (\hat{\beta_0} + \hat{\beta_1} \cdot 1 + \hat{\beta_2} \cdot 1^{2})
$$

Equal to $4.297 - 4.017 = 0.28$ dollars

Considering a discrete change from 10 to 11 years of experience:

- *10 years*: $3.725 + 2.98 - 0.006 (100)$
- *11 years*: $3.725 + 3.278 - 0.006 (121)$
 $\Delta \hat{Y} = 6.277 - 6.105 = 0.172$ dollars



::: {.callout-caution}
## Marginal Effect and Discrete Change

The formula $\Delta\hat{y}/\Delta x \approx \hat{\beta}_1 + 2\hat{\beta}_2 x$ is the **derivative** of the estimated function evaluated at $x$: it measures the change in $\hat{y}$ for an *infinitesimal* change in $x$.


In our example, the marginal effect and the discrete change for $\Delta x = 1$ differ by $\hat{\beta}_2$, which is small in this case ($-0.006$): the marginal effect for a worker with one year of experience is $0.286$, while the discrete change in $wage$ from moving from 1 to 2 years is $0.28$.


When the change in $x$ is not infinitesimal, the effect depends not only on the starting point $x$, but also on the size of the change itself. The difference is equal to $\hat{\beta}_2 \cdot (\Delta x)^2$ and becomes larger as $\Delta x$ increases. In our case, we considered discrete changes of one year, so $\Delta x = 1$ and the difference is $-0.006$, exactly the coefficient on $exper^2$.


In some cases, we prefer the approximation because it provides an **immediate summary** of the effect of $x$ on $y$: it is enough to substitute a value of interest to read the marginal effect at that point. But when the discrete change in $x$ is large, it is more appropriate to calculate the exact change, because the approximation may be misleading.
:::



::: {.callout-note}
## Finding the Peak

When $\hat{\beta_1} > 0$ and $\hat{\beta_2} < 0$, the estimated curve is a **downward-opening parabola**: it first increases, reaches a maximum, and then decreases. This is clear in the previous plot (green point).

The first derivative

$$
\frac{d \hat{y}}{d x} = \hat{\beta_1} + 2 \hat{\beta_2} x
$$

measures the slope of the curve. For small values of $x$, it is **positive** (the curve rises); as $x$ increases, it becomes smaller until it crosses the horizontal axis, that is, until it equals **zero**, exactly at the maximum point. Beyond that threshold, it becomes **negative**: the marginal effect of an additional year of experience on predicted wage changes sign.

Visually, the derivative is a **decreasing line** (because $\hat{\beta_2} < 0$) that crosses the horizontal axis exactly at the peak of the parabola:

```{r}
#| code-fold: true
#| code-summary: "Show code"
b1 <- coef(reg_wage_exper)["exper"]
b2 <- coef(reg_wage_exper)["I(exper^2)"]
xstar <- -b1 / (2 * b2)

deriv_df <- tibble(
  exper = seq(0, 50, length.out = 200),
  d = b1 + 2 * b2 * exper
)

ggplot(deriv_df, aes(x = exper, y = d)) +
  geom_hline(yintercept = 0, color = "grey40", linewidth = 0.4) +
  geom_vline(xintercept = xstar, color = "grey60", linewidth = 0.4, linetype = "dashed") +
  geom_line(color = "#ff7f0e", linewidth = 1.5) +
  annotate("point", x = xstar, y = 0, color = "#2ca02c", size = 5) +
  labs(
    title = "First Derivative: Marginal Effect of One Year of Experience",
    x = "Years of experience",
    y = "Estimated marginal effect"
  ) +
  theme_minimal()
```

To find the peak, set $\frac{d \hat{y}}{d x} = 0$ and solve for $x$:

$$
x^{*} = -\frac{\hat{\beta_1}}{2 \hat{\beta_2}}
$$

In our model:

$$
x^{*} = -\frac{0.298}{2 \cdot (-0.006)} \approx 24.8 \text{ years of experience}
$$

:::



## Logs and Polynomials:

What if $wage$ were expressed in logs? Consider the following regression:

```{r}
reg_logwage_exper <- feols(log(wage) ~ exper + exper^2 + educ + tenure, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_wage_exper, "Log Wage" = reg_logwage_exper), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


**Second model:** The marginal effect when $exper = 0$ is $0.033$, that is, an increase in predicted wage of about 3.3%. Thus, the estimated return to an additional year of experience on $wage$ for a person with no experience is about 3.3%.


The marginal effect at 1 year of experience (for a worker with 1 year of experience) is $0.033 - 2 \cdot 0.001 \cdot 1 = 0.031$, that is, a predicted wage that is **about 3.1% higher**.

The marginal effect at 10 years of experience is $0.033 - 2 \cdot 0.001 \cdot 10 = 0.013$, that is, an increase in predicted wage of about 1.3%.




## Deviation from the Mean

In the previous regression, we saw that the coefficient on $exper$ represents the marginal effect for a worker with zero years of experience. In some cases, it can be useful to consider deviations from the mean. We create a new variable:

$$
exper\_center = exper - mean(exper)
$$


```{r}
wage1 <- wage1 %>%
  mutate(exper_center = exper - mean(exper, na.rm = TRUE))
```


What is the intuition for using deviations from the mean? Consider a simple regression without a polynomial:

```{r}
reg_experc <- feols(wage ~ exper_center, data = wage1, vcov = "hetero")

reg_exper <- feols(wage ~ exper, data = wage1, vcov = "hetero")
modelsummary(list("Wage" = reg_experc, "Wage" = reg_exper), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

What changes?

- **The intercept!** In the first model, it represents the predicted wage for an individual with **experience equal to the mean**.

- In the second model, the intercept represents the predicted wage with zero years of experience.

Thus, expressing $exper$ as a deviation from the mean allows us to interpret the intercept as the predicted wage for an individual with average experience, rather than for an individual with no experience.


## Deviation from the Mean and Polynomials


In polynomial regressions, expressing $exper$ as a deviation from the mean allows us to interpret the coefficient on $exper\_center$ as the marginal increase in wage when $exper$ is equal to its mean, rather than when $exper$ is equal to zero.


```{r}
reg_wagepolcenter <- feols(wage ~ educ + exper_center + exper_center^2 + tenure, data = wage1, vcov = "hetero")
reg_wagepol <- feols(wage ~ educ + exper + exper^2 + tenure, data = wage1, vcov = "hetero")

modelsummary(list("Wage" = reg_wagepolcenter, "Wage" = reg_wagepol), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


Now the coefficient on $exper\_center$ represents the marginal increase in wage when $exper$ is equal to its mean. The coefficients on the other variables (**except the intercept**) remain unchanged.


The coefficient on $exper\_center$, equal to 0.062, represents the increase in wage when $exper$ is equal to its mean, that is, when $exper\_center$ is equal to zero.

By contrast, in the second model (without $exper\_center$), the coefficient on $exper$ represents the marginal increase in wage for a worker with zero years of experience.



Recall that:

$$
exper\_center = exper - mean(exper)
$$


The descriptive statistics for $exper$ are:

```{r}
desc_exper <- wage1 %>%
  summarise(
    Mean = mean(exper, na.rm = TRUE),
    Q1 = quantile(exper, 0.25, na.rm = TRUE),
    Mediana = median(exper, na.rm = TRUE),
    Q3 = quantile(exper, 0.75, na.rm = TRUE)
  )

tt(desc_exper, digits = 3)
```

The mean of $exper$ is the value we subtract to obtain $exper\_center$: by construction, the latter has **mean zero**.



::: {.callout-caution}
## Centering and Multicollinearity

Beyond the interpretive advantage just discussed, centering $exper$ also has a **computational** advantage: it reduces multicollinearity between the linear and quadratic terms.

The intuition is as follows. If $exper$ varies around its mean $\overline{exper}$, we can approximate $exper^2$ with a first-order expansion around $\overline{exper}$:

$$
exper^2 \approx \overline{exper}^{\,2} + 2\,\overline{exper}\,(exper - \overline{exper}) = -\overline{exper}^{\,2} + 2\,\overline{exper}\cdot exper.
$$

When $\overline{exper}$ is far from zero (in our case, $\overline{exper} \approx 17$), $exper^2$ behaves **almost like a linear function of $exper$**, with slope $2\,\overline{exper}$. The two regressors, $exper$ and $exper^2$, therefore tend to move together: for OLS it becomes difficult to separate the contribution of each, standard errors inflate, and numerical computation is less stable.

If instead we use $exper\_center = exper - \overline{exper}$, the same expansion gives $exper\_center^2 \approx 0$ near the mean: even powers ($exper\_center^2$) and odd powers ($exper\_center$) tend to be **orthogonal to each other**. The regressor matrix is better conditioned and standard errors decrease.

The same argument applies to higher-order polynomials ($exper^3$, $exper^4$, ...): centering before raising to powers always improves numerical conditioning.
:::
