---
title: "Introduction to R"
subtitle: "Regression with fixest 📉"
author: "Econometrics"
date: today
lang: en

format:
  revealjs:
    theme: white
    slide-number: c/t
    highlight-style: github
    code-line-numbers: false
    transition: none
    footer: "Econometrics — Sapienza University of Rome"
    embed-resources: true
    width: 1280
    height: 720

execute:
  warning: false
  message: false
  echo: true

knitr:
  opts_chunk:
    comment: ""
    fig.align: "center"
    fig.width: 9
    fig.height: 5
---



```{r}
#| include: false
library(tidyverse)
library(wooldridge)
library(fixest)
library(modelsummary)

data("wage1", package = "wooldridge")
```

## Libraries and data

```{r}
#| eval: false
library(tidyverse)
library(wooldridge)
library(fixest)
library(modelsummary)

data("wage1", package = "wooldridge")
```

- `wooldridge` — library with datasets
- `fixest` — linear regression with robust standard errors
- `modelsummary` — regression tables

---

## 🗂️ Data structure

```{r}
head(wage1)
```

Variables we will use:

- `wage` — average hourly earnings (dollars, 1976)
- `educ` — years of education


---

## 📊 Descriptive statistics

```{r}
wage1 |>
  summarise(
    wage_mean   = mean(wage),
    wage_median = median(wage),
    wage_sd     = sd(wage),
    educ_mean   = mean(educ),
    educ_median = median(educ),
    educ_sd     = sd(educ)
  )
```

---

## 📊 Data visualization
```{r}
#| layout-ncol: 2
#| fig-width: 6
#| fig-height: 4

# Distribution of hourly wage
ggplot(wage1, aes(x = wage)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 20) +
  labs(title = "Distribution of hourly wage",
        x = "Hourly wage (dollars)", y = "Frequency") +
  theme_minimal(base_size = 13)

# Distribution of years of education
ggplot(wage1, aes(x = educ)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 20) +
  labs(title = "Distribution of years of education",
       x = "Years of education", y = "Frequency") +
  theme_minimal(base_size = 13)
```


---

## 🔍 Relationship between wage and educ

- `method = "lm"` — adds a linear regression line

```{r}
#| output-location: slide
ggplot(wage1, aes(y = wage, x = educ)) +
  geom_point(color = "black", alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(x = "Years of education",
       y = "Hourly wage (dollars)") +
  theme_minimal(base_size = 13)
```



---


## 📉 Regression — simple linear model

We estimate the simple linear regression model:

$$wage_i = \beta_0 + \beta_1 \, educ_i + u_i$$

```{r}
#| output-location: slide
reg1 <- feols(wage ~ educ, data = wage1, vcov = "hetero")
reg1
```

- `vcov = "hetero"` or `vcov = "hc1"` — robust standard errors
- Stata command: `regress wage educ, robust`


---

## 📉 Regression — R's `lm` command

- the base command for linear regression is `lm()`, but it does not provide robust standard errors (you need `vcovHC()` from the `sandwich` package)

```{r}
reg_lm <- lm(wage ~ educ, data = wage1)
summary(reg_lm)
```

---

## 📋 Regression table

- dependent variable: `wage`
- robust standard errors in parentheses

::: {style="font-size: 0.85em"}
```{r}
modelsummary(list("wage" = reg1),gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```
:::


---



## 📋 Regression table

The coefficient on $educ$ represents the average increase in hourly wage associated with an additional year of education:

- An additional year of education is **associated** with an average increase in hourly wage of approximately 0.54 dollars


The intercept ($\beta_0$) represents the predicted average hourly wage for an individual with zero years of education
