---
title: "Problem Set N.1"
subtitle: "College GPA"
author: "Andrea Recine"
date: today
lang: en

format:
  html:
    # Layout
    theme: flatly
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: right
    number-sections: true

    # Code
    code-fold: false
    code-tools: true
    code-block-bg: true
    code-block-border-left: false
    highlight-style: github

    # Output
    embed-resources: true

    # Typography
    fontsize: 1.1em
    linestretch: 1.6

execute:
  warning: false
  message: false
  echo: true

knitr:
  opts_chunk:
    comment: "#>"
    fig.align: "center"
    fig.width: 7
    fig.height: 4.5

---

```{r}
#| include: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(tidyverse)
library(dslabs)
library(fixest)
library(modelsummary)
```

## Instructions {.center}

- Complete the exercises using R
- For each exercise, show the code and comment on the results

---


## College GPA:

`gpa1` contains the college grade point average ($colGPA$), high school GPA ($hsGPA$), and achievement test score ($ACT$) of 141 students from large universities.


$colGPA$ and $hsGPA$ are on a four-point scale.

$ACT$ (American College Testing) is a standardized test widely used in the US for college admission. It consists of 4 sections and is scored on a scale from 1 to 36.

```{r}
data("gpa1", package = "wooldridge")
```

```{r}
head(gpa1)
```

```{r}
# Descriptive statistics
statgpa <- gpa1 %>%
       summarise(
       colgpa_mean = mean(colGPA, na.rm = TRUE),
       hsgpa_mean = mean(hsGPA, na.rm = TRUE),
       act_mean = mean(ACT, na.rm = TRUE),
  )

# Display results
statgpa %>%
  kable(caption = "Descriptive statistics",
        digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
summary(gpa1$colGPA)
summary(gpa1$hsGPA)
```

### Exercise 1:

Consider the following regression:
$$
colGPA = \beta_0 + \beta_1 hsGPA_i + \beta_2 ACT_i + u_i
$$

Estimate it using `fixest` and interpret the coefficients on $hsGPA$ and $ACT$.

```{r}
reg_gpa1 <- feols(colGPA ~ hsGPA + ACT, data = gpa1, vcov = "hetero")
modelsummary(list("colGPA" = reg_gpa1), output = "markdown", gof_omit = "AIC|BIC|RMSE")
```

Holding $ACT$ constant, one additional point of high school GPA is associated, on average, with 0.453 additional points of $colGPA$.


**Note**: the maximum value of $hsGPA$ is 4.


$R^2$: $hsGPA$ and $ACT$ together explain 17.6 \% of the variation in $colGPA$.


Let us change the scale of $hsGPA$
```{r}
gpa1 <- gpa1 %>%
  mutate(hsGPA100 = hsGPA *25)
```

```{r}
reg_gpa2 <- feols(colGPA ~ hsGPA100 + ACT, data = gpa1, vcov = "hetero")
modelsummary(list("colGPA100" = reg_gpa2), output = "markdown", gof_omit = "AIC|BIC|RMSE")
```

A 1-point increase (on a scale up to 100) in $hsGPA$ is associated, on average, with an increase of 0.0181 points in $colGPA$.

As before, 0.0181 = 0.453/25

We add as a regressor $skipped$, which represents the average number of lectures missed (skipped) per week:

$$
colGPA = \beta_0 + \beta_1 hsGPA_i + \beta_2 ACT_i + \beta_3 skipped_i + u_i
$$


```{r}
reg_gpa3 <- feols(colGPA ~ hsGPA + ACT + skipped, data = gpa1, vcov = "hetero")
modelsummary(list("colGPA" = reg_gpa1, "colGPA" = reg_gpa3), output ="markdown", gof_omit = "AIC|BIC|RMSE")
```

A 1-point increase in $hsGPA$ is associated, on average, with an increase of 0.412 points in $colGPA$, holding $ACT$ and $skipped$ constant.

What if we were interested in the other two coefficients?

- A 1-point increase in $ACT$, holding $hsGPA$ and $skipped$ constant, is associated with an increase, on average, of 0.015 points in $colGPA$. The t-statistic is 1.36 — what does that mean?

- Holding $hsGPA$ and $ACT$ constant, one additional missed (skipped) lecture per week is associated with a decrease, on average, of 0.083 points in $colGPA$.

### Effects of PC ownership on college GPA

We regress $colGPA$ on a dummy variable $PC$ equal to one if the student owns a personal computer.

We estimate the following regressions:

- `reg_gpa` simple linear regression
- `reg_gpa_pc` regression with $hsGPA$ and $ACT$
- `reg_gpa_pc3` regression of $hsGPA$ on $ACT$, $mothcoll$ (= 1 if the mother has at least a bachelor's degree, BA) and $fathcoll$ (= 1 if the father has at least a bachelor's degree, BA).

```{r}
reg_gpa_pc1 <- feols(colGPA ~ PC, data = gpa1, vcov = "hetero")
reg_gpa_pc2 <- feols(colGPA ~ PC + hsGPA + ACT , data = gpa1, vcov = "hetero")
reg_gpa_pc3 <- feols(colGPA ~ PC + hsGPA + ACT + mothcoll + fathcoll , data = gpa1, vcov = "hetero")
modelsummary(list("colGPA" = reg_gpa_pc1, "colGPA" = reg_gpa_pc2, "colGPA" = reg_gpa_pc3), output = "markdown", gof_omit = "AIC|BIC|RMSE")

```

- First regression: A student who owns a PC has, on average, a $colGPA$ about 0.17 points higher than a student who does not own one

- Second regression: Holding $hsGPA$ and the $ACT$ score constant, a student who owns a PC has, on average, a $colGPA$ about 0.16 points higher than a student who does not own one.

- Third regression: Holding $hsGPA$, the $ACT$ score, and parental education ($mothcoll$ and $fathcoll$) constant, a student who owns a PC has, on average, a $colGPA$ about 0.15 points higher than a student who does not own one.










## Data from `dslabs`

The `gapminder` dataset contains data on health and economic development by country and year.

```{r}
library(dslabs)
head(gapminder)
```

Main variables:

- `country` — country
- `year` — year
- `life_expectancy` — life expectancy (years)
- `gdp` — total GDP (dollars)
- `population` — population
- `continent` — continent

---

## Sample Selection

We work with a single year to avoid the panel dimension. We filter to **2011** and remove missing observations:

```{r}
gap2011 <- gapminder |>
  filter(year == 2011) |>
  filter(!is.na(gdp), !is.na(life_expectancy), !is.na(population)) |>
  mutate(gdp_pc = gdp / population)   # GDP per capita
```

- `filter()` selects the rows that satisfy a condition
- `!is.na()` removes missing values
- `mutate()` creates new variables

---

## Exercise 1 — Descriptive Statistics

Compute the mean, median, and standard deviation of `life_expectancy` and `gdp_pc`:

```{r}
#| eval: false
gap2011 |>
  summarise(
    le_mean    = mean(...),
    le_median  = median(...),
    le_sd      = sd(...),
    gdp_mean   = mean(...),
    gdp_median = median(...),
    gdp_sd     = sd(...)
  )
```

---

## Exercise 2 — Statistics by Continent

Compute the mean and median of `life_expectancy` and `gdp_pc` **by continent**:

```{r}
#| eval: false
gap2011 |>
  group_by(...) |>
  summarise(
    le_mean    = mean(...),
    le_median  = median(...),
    gdp_mean   = mean(...),
    gdp_median = median(...)
  )
```

- `group_by()` groups the data by a categorical variable before `summarise()`

---

## Exercise 3 — Distribution of `life_expectancy`

Construct a histogram of `life_expectancy` and a boxplot by continent:

```{r}
#| eval: false
# Histogram
ggplot(gap2011, aes(x = life_expectancy)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 20) +
  labs(title = "Distribution of life expectancy (2011)",
       x = "Life expectancy (years)", y = "Frequency") +
  theme_minimal()

# Boxplot by continent
ggplot(gap2011, aes(x = ..., y = life_expectancy, fill = continent)) +
  geom_boxplot() +
  labs(title = "Life expectancy by continent (2011)",
       x = "Continent", y = "Life expectancy (years)") +
  theme_minimal()
```

---

## Exercise 4 — Scatterplot

Construct a scatterplot of `life_expectancy` (y-axis) against `gdp_pc` (x-axis). Color the points by continent.

What do you observe? Does the relationship appear linear?

```{r}
#| eval: false
ggplot(gap2011, aes(x = gdp_pc, y = life_expectancy, color = continent)) +
  geom_point(...) +
  labs(title = "GDP per capita and life expectancy (2011)",
       x = "GDP per capita", y = "Life expectancy (years)") +
  theme_minimal()
```

---

## Exercise 5 — Logarithmic Scale

Repeat the previous plot using `log(gdp_pc)` on the x-axis:

```{r}
#| eval: false
ggplot(gap2011, aes(x = log(gdp_pc), y = life_expectancy, color = continent)) +
  geom_point(...) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.5) +
  labs(title = "Log GDP per capita and life expectancy (2011)",
       x = "Log GDP per capita", y = "Life expectancy (years)") +
  theme_minimal()
```

How does the relationship change compared to the previous plot?

---

## Exercise 6 — Baseline Regression

Estimate the following regression with robust standard errors:

$$life\_expectancy_i = \beta_0 + \beta_1 \log(gdp\_pc_i) + u_i$$

```{r}
#| eval: false
gap2011 <- gap2011 |>
  mutate(log_gdp_pc = log(gdp_pc))

reg1 <- feols(life_expectancy ~ log_gdp_pc, data = gap2011, vcov = "hetero")
reg1
```

**Questions:**

1. How should $\hat{\beta}_1$ be interpreted? *(careful: $x$ is in logs)*
2. Is the coefficient statistically significant at the 5% level?
3. What is the value of $R^2$?

---

## Exercise 7 — Continent Dummies

Add `continent` as a regressor:

$$life\_expectancy_i = \beta_0 + \beta_1 \log(gdp\_pc_i) + \sum_{j} \gamma_j \, continent_{ij} + u_i$$

```{r}
#| eval: false
reg2 <- feols(life_expectancy ~ log_gdp_pc + continent,
              data = gap2011, vcov = "hetero")

modelsummary(list("Base" = reg1, "With continent" = reg2),
             gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

**Questions:**

1. What is the reference category (*base category*)?
2. How does $\hat{\beta}_1$ change relative to the baseline model? Why?
3. How does $R^2$ change?
