---
title: "Review of log models"
author: "Andrea Recine"
---

**Required packages:**

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(modelsummary)
library(fixest)
```

**For the data:**
```{r}
#| warning: false
library(wooldridge)
```

```{r}
data("wage1", package = "wooldridge")
```


::: {.callout-note title="Log specifications"}

| Model form              | Interpretation                                                                     |
|-------------------------|----------------------------------------------------------------------------------|
| $\log(Y)$ ~ $X$         | A one-unit increase in $X$ leads to a change of $100 \cdot \beta_1$\% in $Y$              |
| $Y$ ~ $\log(X)$         | A 1\% increase in $X$ leads to a change of $0.01 \cdot \beta_1$ in $Y$       |
| $\log(Y)$ ~ $\log(X)$   | A 1\% increase in $X$ leads to a change of $\beta_1$\% in $Y$                    |

:::


**3 cases with wage and educ:**

```{r}
reg_logwage <- feols(log(wage) ~ educ, data = wage1, vcov = "hetero")
reg_wagelog <- feols(wage ~ log(educ), data = wage1, vcov = "hetero")
reg_logwagelog <- feols(log(wage) ~ log(educ), data = wage1, vcov = "hetero")
modelsummary(list("Log Wage" = reg_logwage, "Wage" = reg_wagelog, "Log Wage" =reg_logwagelog), output = "markdown", gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


- **First model** (log-level): Each additional year of education is associated, on average, with an 8.3% increase in the hourly wage.

- **Second model** (level-log): A 1% increase in years of education is associated, on average, with an increase in the hourly wage of 0.053 dollars. Equivalently, a 10% increase in years of education is associated, on average, with an increase in the hourly wage of 0.53$ (10 $\cdot$ 0.053).

- **Third model** (log-log): A 1% increase in years of education is associated, on average, with a 0.825% increase in the hourly wage.

The choice of specification (functional form) should be guided by reasoning: which interpretation makes the most sense?

Logarithms are often used when the variable is measured in dollars (for example, wage or income) or for variables such as population (particularly when there is enough variation).

They are less commonly used for variables measured in years or that are already percentages or proportions (for example, the unemployment rate).


::: {.callout-note title="Units of measurement of the dependent variable"}

With variables expressed in logs, changing the unit of measurement is irrelevant. Try expressing wage in hundreds of dollars by creating $wage$/100 and running the regression on $educ$.
The same holds if $X$ and $Y$ are both in logs.
The intercept, however, will change.
:::



::: {.callout-note title="Confidence intervals"}

Compute the 95% confidence interval for the coefficient on $educ$ in the Log Wage regression.

What is the confidence interval for 10 additional years of education? The upper and lower endpoints will equal the previous ones multiplied by 10. Try it.
:::


**Change in $wage$**

```{r}
reg_logwage100 <- feols(log(wage/100) ~ educ, data = wage1, vcov = "hetero")
reg_wagelog100 <- feols(wage/100 ~ log(educ), data = wage1, vcov = "hetero")
reg_logwagelog100 <- feols(log(wage/100) ~ log(educ), data = wage1, vcov = "hetero")
modelsummary(list("Log Wage" = reg_logwage100, "Wage" = reg_wagelog100, "Log Wage" =reg_logwagelog100), output = "markdown", gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

- In the first model, the intercept changes
- In the second model, a 1% increase in years of education increases $wage$ by (0.053/100) hundreds of dollars
- In the third model, only the intercept changes


**Change in $educ$:**

Let us express years of education in months of education:


```{r}
wage1$educ12 <- wage1$educ * 12
```

```{r}
reg_logwage12 <- feols(log(wage) ~ educ12, data = wage1, vcov = "hetero")
reg_wagelog12 <- feols(wage ~ log(educ12), data = wage1, vcov = "hetero")
reg_logwagelog12 <- feols(log(wage) ~ log(educ12), data = wage1, vcov = "hetero")

modelsummary(list("Log Wage" = reg_logwage12, "Wage" = reg_wagelog12, "Log Wage" =reg_logwagelog12), output = "markdown", gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

What has changed?

- An additional month of education is associated with an increase in the hourly wage of approximately 0.7%
- A 1% increase in months of education is associated with an increase in the hourly wage of 5.330/100 dollars
- A 1% increase in months of education is associated with a 0.825% increase in the wage


## With other variables:

```{r}
reg_logwage_mul <- feols(log(wage) ~ educ + exper + tenure + female, data = wage1, vcov = "hetero")
reg_wagelog_mul <- feols(wage ~ log(educ) + exper + tenure + female, data = wage1, vcov = "hetero")
reg_logwagelog_mul <- feols(log(wage) ~ log(educ) + exper + tenure + female, data = wage1, vcov = "hetero")
modelsummary(list("Log Wage" = reg_logwage_mul, "Wage" = reg_wagelog_mul, "Log Wage" =reg_logwagelog_mul), output = "markdown", gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```





**Change in $wage$**

```{r}
reg_logwage100_mul <- feols(log(wage/100) ~ educ + exper + tenure + female, data = wage1, vcov = "hetero")
reg_wagelog100_mul <- feols(wage/100 ~ log(educ) + exper + tenure + female, data = wage1, vcov = "hetero")
reg_logwagelog100_mul <- feols(log(wage/100) ~ log(educ) + exper + tenure + female, data = wage1, vcov = "hetero")
modelsummary(list("Log Wage" = reg_logwage100_mul, "Wage" = reg_wagelog100_mul, "Log Wage" =reg_logwagelog100_mul), output = "markdown", gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```



**Change in $educ$:**

Let us express years of education in months of education:


```{r}
wage1$educ12 <- wage1$educ * 12
```

```{r}
reg_logwage12_mul <- feols(log(wage) ~ educ12 + exper + tenure + female, data = wage1, vcov = "hetero")
reg_wagelog12_mul <- feols(wage ~ log(educ12) + exper + tenure + female, data = wage1, vcov = "hetero")
reg_logwagelog12_mul <- feols(log(wage) ~ log(educ12) + exper + tenure + female, data = wage1, vcov = "hetero")

modelsummary(list("Log Wage" = reg_logwage12_mul, "Wage" = reg_wagelog12_mul, "Log Wage" =reg_logwagelog12_mul), output = "markdown", gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


::: {.callout-note title="Exercise"}

Compare the regressions
:::