---
title: "Problem Set 1"
subtitle: "Questions"
author: "Andrea Recine"
date: today
lang: en
---

# Problem Set 1


**Libraries to load**:

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(tinytable)
library(modelsummary)
library(fixest)
library(car)
library(marginaleffects)
```


```{r}
#| warning: false
library(wooldridge)
```



## Exercise 1

Consider the data in `wage2`

```{r}
data("wage2", package = "wooldridge")
```


- $wage$: monthly earnings
- $educ$: years of education
- $exper$: years of experience
- $tenure$: years with the current employer
- $black$: dummy variable indicating whether the worker is Black



We estimate the following models:

```{r}
modelA <- feols(wage ~ educ + exper + tenure + black, data = wage2, vcov = "hetero")
```


```{r}
modelB <- feols(log(wage) ~ educ + exper + tenure + black, data = wage2, vcov = "hetero")
```



```{r}
modelsummary(list("Wage (Model A)" = modelA, "Log Wage (Model B)" = modelB), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


### Questions

::: {.callout-tip title="Question 1 — Interpretation of `educ`"}
Interpret the coefficient on $educ$ in **Model A** (level) and in **Model B** (log).
:::



::: {.callout-tip title="Question 2 — Change of units of measurement"}
Suppose that the wage $wage$ in **Model A** is expressed in hundreds of dollars (that is, `wage/100`). Which of the following statements is correct?

a) The coefficients of Model A would all be divided by 100, while those of Model B would remain unchanged.

b) The coefficients of Model A would all be divided by 100, while those of Model B, except the intercept, would remain unchanged.

c) The coefficients of Model A would all be multiplied by 100, while those of Model B, except the one on `educ`, would remain unchanged.

d) The coefficients of Model A would all be multiplied by 100, while those of Model B would remain unchanged.

:::




::: {.callout-tip title="Question 3 — Coefficient on `black` (multiple choice)"}
Which interpretation of the coefficient on $black$ is correct?

a) In Model A: Black workers earn on average about 175.13 dollars less per month, holding the other variables constant.
b) In Model B: Black workers earn on average about 20.7% less, holding the same variables constant.
c) Both of the above, one for each model.
:::



::: {.callout-tip title="Question 4 — 95% CI (calculation)"}
Construct the 95% confidence interval for the coefficient on $educ$ in **Model B**.

:::



::: {.callout-tip title="Question 5 — 90% CI"}
Construct the 90% confidence interval for the coefficient of an increase of 20 years of $educ$ in **Model A**.
:::




::: {.callout-tip title="Question 6 — F test of joint significance"}
Test the joint null hypothesis $H_0: \beta_{exper} = 0 \;\text{and}\; \beta_{tenure} = 0$ in **Model B**. Can we **reject/not reject** $H_0$ at the 5% level?
:::

```{r}
#| code-fold: true
linearHypothesis(modelB, c("exper = 0", "tenure = 0"))
```


## Exercise 2


We use the `wage1` dataset (Wooldridge), a sample of US workers from the 1976 CPS survey. Warning: in this section **experience is measured in decades** (`exper_dec = exper / 10`).

```{r}
data("wage1", package = "wooldridge")

wage1 <- wage1 |>
  mutate(exper_dec  = exper / 10,   # experience in decades
         exper_dec2 = exper_dec^2)  # experience (in decades) squared
```

| variable    | description |
|-------------|-------------|
| `wage`      | hourly wage in dollars |
| `educ`      | years of education completed |
| `exper_dec` | decades of work experience |
| `tenure`    | years with the current employer |
| `female`    | =1 if woman, =0 if man |
| `married`   | =1 if married, =0 otherwise |
| `nonwhite`  | =1 if nonwhite, =0 otherwise |

The researchers estimate six specifications:

```{r}
mA <- feols(wage      ~ educ + exper_dec + tenure + female,
            data = wage1, vcov = "hetero")                                   # A: linear

mB <- feols(log(wage) ~ educ + exper_dec + tenure + female,
            data = wage1, vcov = "hetero")                                   # B: log-linear

mC <- feols(log(wage) ~ log(educ) + exper_dec + tenure + female,
            data = wage1, vcov = "hetero")                                   # C: log-log

mD <- feols(log(wage) ~ educ + exper_dec + female:exper_dec + tenure + female,
            data = wage1, vcov = "hetero")                                   # D: interaction

mE <- feols(log(wage) ~ educ + exper_dec + exper_dec2 + tenure + female,
            data = wage1, vcov = "hetero")                                   # E: polynomial

mF <- feols(log(wage) ~ educ + educ:female + exper_dec + exper_dec2 + tenure + female,
            data = wage1, vcov = "hetero")                                   # F: interaction + polynomial

modelsummary(list("(A)" = mA, "(B)" = mB, "(C)" = mC,
                  "(D)" = mD, "(E)" = mE, "(F)" = mF),
             gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

All models report heteroskedasticity-robust standard errors.

### Questions

::: {.callout-tip title="Question 1 — Models A and B"}
In **Model A**, an additional year of education is associated with an increase in the hourly wage of about **__________** *(dollars / %)*.

In **Model B**, an additional year of education is associated with an increase in the hourly wage of about **__________** *(dollars / %)*.
:::



::: {.callout-tip title="Question 2 — Change of units of measurement"}
If we re-estimated the models expressing the wage in **tens of dollars** (`wage/10`) instead of dollars, indicate **all** the correct statements:

1. The $R^2$ values would be different.
2. In Model A the standard errors would be multiplied by 10.
3. Statistical inference (the significance of the coefficients) would remain unchanged.
4. In the log-linear model the intercept would change, but not the coefficients of the explanatory variables.
5. Rescaling the dependent variable by a positive constant does not change the coefficients of the explanatory variables in the log-linear model.
6. The coefficients would all be multiplied by 10.

Moreover, in **Model A** the coefficient on education would be multiplied by **__________**.
:::

::: {.callout-tip title="Question 3 — Model D (interaction)"}
In **Model D**, the effect of an additional decade of experience (`exper_dec`) for a **woman** is an increase of about **__________** %.

Indicate **all** the correct statements about the interaction `female:exper_dec` ($female \times exper\_dec$):

1. The return to experience on the wage differs between men and women.
2. The coefficient on `female:exper_dec` indicates how much the return to experience for women differs from that for men.
3. The logarithmic transformation captures any nonlinearity anyway, so the interaction is superfluous.
4. For a man (`female = 0`), an additional decade of experience is associated with a change equal to the coefficient on `exper_dec`.
5. The return to experience is identical for men and women.
:::

::: {.callout-tip title="Question 4 — Model E (polynomial)"}
In **Model E**, the marginal effect of experience on the (log) wage for a worker with **10 years of experience** (`exper_dec = 1`) is about **__________** /%.

The number of years of experience at which the marginal effect is zero is **__________** years.

Based on the $t$ test on the coefficient of `exper_dec²`, at the 10% level (**we reject / we do not reject**) **__________** the null hypothesis that the relationship between the wage and experience, holding other conditions constant, is linear.
:::

::: {.callout-tip title="Question 5 — Comparing the R² values"}
Compare the $R^2$ of Model A with that of Model B. A student concludes that Model B fits the data better because it has a higher $R^2$. Which statement is correct?

a) The comparison is not valid because the two models have different dependent variables (`wage` vs `log(wage)`): the $R^2$ values measure shares of variance of different quantities.
b) The comparison is not valid because Model B has more regressors.
c) The comparison is valid only if the coefficients are statistically significant.
d) The comparison is valid: a higher $R^2$ always indicates a better fit.
:::

::: {.callout-tip title="Question 6 — Model F (interaction + polynomial)"}
In **Model F**, the percentage return to an additional year of education for **men** is about **__________** %, while for **women** it is about **__________** %.

The difference in absolute value between the two returns is about **__________** percentage points.

:::


## Exercise 3

Consider the `hprice1` data (88 houses, Wooldridge), which contain information on house prices and their characteristics.

```{r}
data("hprice1", package = "wooldridge")
```

- $price$: house price in thousands of dollars
- $sqrft$: size in square feet
- $lotsize$: lot size (square feet)
- $bdrms$: number of bedrooms
- $colonial$: dummy = 1 if the house is colonial style

We estimate two models, one **lin-lin** and one **log-log**:

```{r}
model_lin <- feols(price ~ sqrft + lotsize + bdrms + colonial,
                     data = hprice1, vcov = "hetero")

model_log <- feols(log(price) ~ log(sqrft) + log(lotsize) + bdrms + colonial,
                     data = hprice1, vcov = "hetero")

modelsummary(list("Model A" = model_lin, "Model B" = model_log),
             gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


### Questions

::: {.callout-tip title="Question 1 — Elasticity"}
Interpret the coefficient on $\log(sqrft)$ in the log-log model. Why is it called an **elasticity**?
:::



::: {.callout-tip title="Question 2 — Coefficient on `bdrms` (multiple choice)"}
In the log-log model, the coefficient on $bdrms$ ($\hat\beta = 0.027$) is interpreted as:

a) The elasticity of $price$ with respect to $bdrms$.
b) The percentage change in $price$ for one additional bedroom (semi-elasticity).
c) The change in $price$, in dollars, for one additional bedroom.
d) The change in $price$, in thousands of dollars, for one additional bedroom.
:::





::: {.callout-tip title="Question 3 — Coefficient on `colonial`"}
Interpret the coefficient on $colonial$ in the **log-log** model. Approximately how large is the percentage price difference between colonial and non-colonial houses? In the **lin-lin** model, instead, how large is the difference and in what unit of measurement?
:::



::: {.callout-tip title="Question 4 — Change of units of measurement (T/F)"}
If we changed the unit of measurement of $price$ from thousands of dollars ($\$1000$) to dollars ($\$1$), the coefficient on $\log(sqrft)$ would remain unchanged. True or false?
:::



## Exercise 4

Consider the `mlb1` data (353 Major League Baseball players, Wooldridge), which contain data on salaries and career statistics.

```{r}
data("mlb1", package = "wooldridge")
```

- $salary$: annual salary in dollars
- $years$: years in MLB
- $bavg$: batting average (hits per 1000 at bats)
- $hrunsyr$: home runs per year
- $rbisyr$: runs batted in (RBI) per year

We estimate a model with $years$ in **quadratic** form:

$$\log(salary) = \beta_0 + \beta_1 \, years + \beta_2 \, years^2 + \beta_3 \, bavg + \beta_4 \, hrunsyr + u$$

```{r}
model_quad <- feols(log(salary) ~ years + I(years^2) + bavg + hrunsyr,
                      data = mlb1, vcov = "hetero")

modelsummary(list("Model with years²" = model_quad),
             gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


### Questions

::: {.callout-tip title="Question 1 — Why `years²`?"}
Why did we add $years^2$ to the regressors? What does the **sign** of $\hat\beta_2$ (positive or negative) measure?
:::



::: {.callout-tip title="Question 2 — Marginal effect (calculation)"}
Compute the marginal effect of $years$ for a player with 5 years of career and for a player with 15 years of career.
:::






::: {.callout-tip title="Question 3 — If we omit `years²` (multiple choice)"}
If we instead estimated the model **without** $years^2$, the coefficient $\hat\beta_1$ on $years$ would represent:

a) The correct marginal effect (equal to that of the quadratic model).
b) A linear approximation of a relationship that is actually nonlinear; the sign could be misleading depending on the distribution of $years$.
c) Always zero by construction.
d) The semi-elasticity at $years = 0$.
:::



::: {.callout-tip title="Question 4 — Coefficient on `bavg`"}
A batting average $bavg$ higher by 10 points is associated with how many percentage points more salary, holding $years$ and $hrunsyr$ constant?
:::



## Exercise 5

We stay with the `mlb1` data, now interested in a possible **interaction** between $years$ and the dummy $nl$ (= 1 if the player plays in the National League, = 0 if in the American League):

$$\log(salary) = \beta_0 + \beta_1 \, years + \beta_2 \, nl + \beta_3 \, (years \cdot nl) + \beta_4 \, hrunsyr + u$$

```{r}
model_int <- feols(log(salary) ~ years + nl + years:nl + hrunsyr,
                     data = mlb1, vcov = "hetero")

modelsummary(list("Model with interaction" = model_int),
             gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


### Questions

::: {.callout-tip title="Question 1 — Interpretation of $\\hat\\beta_2$ and $\\hat\\beta_3$"}
How do we interpret the coefficient on $nl$ ($\hat\beta_2$) and the coefficient on the interaction $years \times nl$ ($\hat\beta_3$)?
:::



::: {.callout-tip title="Question 2 — Marginal effect of `years` by league (calculation)"}
What is the marginal effect of $years$ for players in the **AL** ($nl = 0$) and in the **NL** ($nl = 1$)? What are their numerical values?
:::



::: {.callout-tip title="Question 3 — Significance of the interaction (multiple choice)"}
Is the interaction coefficient statistically significant at the 5% level? What can we conclude?

a) The return to experience is zero.
b) The return to experience does **not** differ significantly between AL and NL.
c) The variable $nl$ has no effect on salary.
d) The model is misspecified.
:::



::: {.callout-tip title="Question 4 — Centering `years`"}
We center $years$ at the **sample mean** ($\overline{years} \approx 6.33$) and re-estimate the model. How does the interpretation of the coefficient on $nl$ change?
:::

```{r}
mlb1 <- mlb1 |>
  mutate(years_c = years - mean(years))

model_int_c <- feols(log(salary) ~ years_c + nl + years_c:nl + hrunsyr,
                       data = mlb1, vcov = "hetero")

modelsummary(list("Original" = model_int, "Centered on years" = model_int_c),
             gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```





## Exercise 6 — Binary Dependent Variable (LPM vs Logit)

The labor force participation of married women is a classic topic in labor economics. When the dependent variable is binary (the woman is in the labor force: yes/no), we must choose among different econometric models.

We use the `mroz` dataset (Mroz, 1987), which contains information on 753 married women. We analyze the factors associated with the probability of participating in the labor market.

| variable | description |
|-----------|-------------|
| `inlf`     | =1 if the woman is in the labor force, =0 otherwise |
| `nwifeinc` | family income not earned by the woman (in thousands of \$) |
| `educ`     | years of education |
| `exper`    | years of work experience |
| `expersq`  | experience squared (`exper^2`) |
| `age`      | age in years |
| `kidslt6`  | number of children under 6 years old |
| `kidsge6`  | number of children between 6 and 18 years old |

We estimate two models:

- **Model A** — Linear Probability Model (LPM)
- **Model B** — Logit model

```{r}
data("mroz", package = "wooldridge")

mA_lpm <- feols(inlf ~ nwifeinc + educ + exper + expersq + age + kidslt6 + kidsge6,
                vcov = "hetero", data = mroz)

mB_logit <- feglm(inlf ~ nwifeinc + educ + exper + expersq + age + kidslt6 + kidsge6,
                  family = binomial(link = "logit"), data = mroz)

modelsummary(list("LPM" = mA_lpm, "Logit" = mB_logit), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


The **average marginal effects** (AME) of the Logit model are:

```{r}
marginaleffects::avg_slopes(mB_logit)
```


::: {.callout-tip title="Question 1 — Interpreting the LPM"}
According to the LPM (Model A), having **one additional child under 6 years old** (`kidslt6`) is associated with a (**decrease** / **increase**) of ______ percentage points in the probability of participating in the labor market, holding the other variables constant.
:::



::: {.callout-tip title="Question 2 — Logit coefficients (T/F)"}
Select **all** the correct statements:

1. Logit coefficients represent the effect on the *log(odds)*, not on the probability.
2. In the logit, the marginal effect of a variable on the probability depends on the level of **all** the explanatory variables.
3. One way to interpret them in terms of probability is to compute the average marginal effects (AME).
4. Logit coefficients directly represent the change in the probability.
5. Logit coefficients are always larger (in absolute value) than those of the LPM.
6. In both models the marginal effect is constant for all individuals.
7. Logit coefficients become probabilities if multiplied by 100.
:::




::: {.callout-tip title="Question 3 — Marginal effect in the logit"}
According to Model B (Logit), **4 additional years of education** (`educ`) are associated with a (**decrease** / **increase**) of ______ percentage points in the probability of participating.
:::




::: {.callout-tip title="Question 4 — Confidence interval"}
Still for the effect of **4 additional years of education** in the logit, the 95% confidence interval (in percentage points) is approximately ( ______ , ______ ).
:::




::: {.callout-tip title="Question 5 — LPM vs Logit comparison"}
Indicate **all** the correct statements:

1. In the LPM, **2 additional children** under age 6 are associated with a reduction in the probability of about 52.36 percentage points.
2. In the logit (AME), **one additional child** under age 6 is associated with a reduction of about 25.78 percentage points.
3. In the LPM, an additional year of education is associated with an increase of about 3.80 percentage points.
4. The estimated effects on the probability are identical in the two models.
5. The logit always predicts larger effects than the LPM.
6. Education does not have a statistically significant effect in either model.
:::



::: {.callout-tip title="Question 6 — Difference in predicted probability (LPM)"}
Using the LPM (Model A), compare two women identical in `nwifeinc` (=20), `exper` (=10), `age` (=40), and `kidsge6` (=1), but different as follows:

- **Woman A**: `educ` = 16, `kidslt6` = 0
- **Woman B**: `educ` = 12, `kidslt6` = 2

The difference in the predicted probability of participation (A − B) is about ______ percentage points.
:::


## Exercise 7 — Regression with Panel Data

Do firms that receive **job training grants** really improve their productivity? We measure productivity with the *scrap rate* (`scrap`: defective items per 100 produced); we use its logarithm `lscrap`.

The problem is the **endogeneity** of `grant`: grants are not randomly assigned, and firms with more quality problems may be more likely to apply for them (or to obtain them).


We use the panel dataset `jtrain`, which follows a sample of Michigan manufacturing firms for 3 years (1987–1989).

| variable | description |
|-----------|-------------|
| `fcode`    | firm identification code |
| `year`     | year of the observation |
| `lscrap`   | logarithm of the scrap rate (inverse proxy for productivity) |
| `grant`    | =1 if the firm received the training grant in that year (time-varying) |
| `lsales`   | log of sales |
| `lemploy`  | log of the number of employees |
| `union`    | =1 if the firm has a unionized workforce (time-invariant) |

- **Model A (Pooled OLS)**: ignores the panel structure
- **Model B (firm FE)**: firm fixed effects
- **Model C (firm + year FE)**: firm and year fixed effects

```{r}
data("jtrain", package = "wooldridge")

m_ols <- feols(lscrap ~ grant + lsales + lemploy + union,
               vcov = "hetero", data = jtrain)

m_fe  <- feols(lscrap ~ grant + lsales + lemploy + union | fcode,
               vcov = "cluster", data = jtrain)

m_fet <- feols(lscrap ~ grant + lsales + lemploy + union | fcode + year,
               vcov = "cluster", data = jtrain)

modelsummary(list("Pooled OLS (A)" = m_ols, "Firm FE (B)" = m_fe, "Firm+Year FE (C)" = m_fet), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```




::: {.callout-tip title="Question 1 — Interpreting the Pooled OLS"}
According to Model A (Pooled OLS), receiving the grant in the current year (`grant`) is associated with a scrap rate higher by about ______ (**percentage points** / **dollars** / **%**).
:::



::: {.callout-tip title="Question 2 — Interpreting the fixed effects"}
According to Model B (firm FE), which controls for all fixed characteristics of the firm, receiving the grant in the current year is associated with a scrap rate (**lower** / **higher**) by about ______ (**percentage points** / **dollars** / **%**).
:::




::: {.callout-tip title="Question 3 — Bias and omitted variables"}
Which statements about the bias in the estimate of the grant effect are correct? (multiple selection)

1. The difference between the OLS coefficient and the FE coefficients is too small to be relevant.
2. Firm fixed effects (B and C) correct the bias from any unobserved omitted variable that is constant over time for each firm.
3. Introducing time and firm fixed effects corrects **every** type of endogeneity, including reverse causality and measurement error.
4. Time fixed effects (C) capture shocks common to all firms in a given year (e.g., a recession).
5. Fixed effects cannot do anything about omitted variables that vary over time within a single firm.
:::


::: {.callout-tip title="Question 4 — Which coefficients can be estimated?"}
In Models B and C the coefficient on `union` does not appear. Which statements are correct? (multiple selection)

1. In Models B and C, `union` cannot be estimated because it is **constant over time** for each firm: its variability is entirely absorbed by the firm fixed effects.
2. `grant` can be estimated in the FE models because it varies over time within the same firm.
3. `union` could be estimated in Model C if it were statistically significant.
4. The impossibility of estimating `union` is due to a software error.
5. With firm fixed effects, a time-invariant regressor is perfectly collinear with the fixed effects.
:::


::: {.callout-tip title="Question 5 — Clustered standard errors"}
In Models B and C the standard errors are *clustered* by firm (`fcode`). Which statements are correct? (multiple selection)

1. Clustering is needed because observations from the same firm in different years are not independent (serial correlation in the errors).
2. Using standard errors that are too small increases the probability of wrongly rejecting $H_0$ (Type I error).
3. Omitting clustering, in the presence of serial correlation, tends to produce underestimated standard errors.
4. Clustering corrects the bias in the coefficients due to omitted variables.
5. If a coefficient is not significant, clustering can make it significant.
:::


