---
title: "Problem Set 1"
subtitle: "Current Population Survey (CPS) 2024"
author: "Andrea Recine"
date: today
lang: en

format:
  html:
    # Layout
    theme: flatly
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: right
    number-sections: true

    # Code
    code-fold: false
    code-tools: true
    code-copy: true
    code-block-bg: true
    code-block-border-left: false
    highlight-style: github

    # Output
    embed-resources: true

    # Typography
    fontsize: 1.1em
    linestretch: 1.6

execute:
  warning: false
  message: false
  echo: true

knitr:
  opts_chunk:
    comment: "#>"
    fig.align: "center"
    fig.width: 7
    fig.height: 4.5

---

## Introduction

### Data

The data used in this tutorial come from the **Current Population Survey Annual Social and Economic Supplement (CPS-ASEC)** for the year 2024, extracted via IPUMS CPS [Flood et al., 2025]. The sample is a subselection of 2000 observations of working-age workers (16-65 years old) with positive labor income.

The IPUMS microdata are subject to a use license and are not redistributed. Students interested in replicating the analyses can register for free at <https://cps.ipums.org/> and download a similar extract.

**Full citation**:

> Sarah Flood, Miriam King, Renae Rodgers, Steven Ruggles, J. Robert Warren,
> Daniel Backman, Etienne Breton, Grace Cooper, Julia A. Rivera Drew,
> Stephanie Richards, David Van Riper, and Kari C.W. Williams. *IPUMS CPS:
> Version 13.0* [dataset]. Minneapolis, MN: IPUMS, 2025.
> <https://doi.org/10.18128/D030.V13.0>


```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(tinytable)
library(modelsummary)
library(fixest)
```


```{r}
#| warning: false
library(wooldridge)
```

```{r}
data("wage1", package = "wooldridge")
```





## Exercise 2

Consider the data in `wage2`

```{r}
data("wage2", package = "wooldridge")
```


- $wage$: monthly earnings
- $educ$: years of education
- $exper$: years of experience

We estimate the following regressions:

```{r}
reg_wage1 <- feols(wage ~ educ + exper + tenure, data = wage2, vcov = "hetero")
reg_logwage1 <- feols(log(wage) ~ educ + exper + tenure, data = wage2, vcov = "hetero")

modelsummary(list("Wage" = reg_wage1, "Log Wage" = reg_logwage1), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


::: {.callout-note title="Questions"}
1. How should the coefficients on $educ$ in the two regressions be interpreted?
2. Are the coefficients statistically significant at the 5% level? (Also compute the 95% confidence interval)
3. Wald test (heteroskedasticity-robust F statistic) that all the regression coefficients are equal to zero
:::





### Parental education: dataset `wage2`

We use the `wage2` dataset, considering $pareduc$, which represents the sum of the parents' years of education. $wage$ in `wage2` is the wage expressed in monthly dollars:

```{r}
wage2 <- wage2 %>%
  mutate(pareduc = meduc + feduc)
```

```{r}
regeduc_int <- feols(wage ~ educ*pareduc + exper + tenure, data = wage2, vcov = "hetero")
regeduc_intlog <- feols(log(wage) ~ educ*pareduc + exper + tenure, data = wage2, vcov = "hetero")
modelsummary(list("Wage " = regeduc_int, "Log Wage " = regeduc_intlog), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```




## Deviation from the Mean



```{r}
wage2 <- wage2 %>%
  mutate(
    pareduc_center = pareduc - mean(pareduc, na.rm = TRUE),
    educ_center = educ - mean(educ, na.rm = TRUE)
  )

```


Comparison with the previous regression:

```{r}
regeduc_int <- feols(wage ~ educ*pareduc + exper + tenure, data = wage2, vcov = "hetero")
regeduc_int_mean <- feols(wage ~ educ_center * pareduc_center + exper + tenure, data = wage2, vcov = "hetero")
modelsummary(list("Wage " = regeduc_int, "Wage " = regeduc_int_mean), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```












**References**

**Data source**: The data used come from the CPS ASEC 2024,
extracted via IPUMS CPS:

Flood, S., King, M., Rodgers, R., Ruggles, S., Warren, J. R., Backman, D.,
Breton, E., Cooper, G., Rivera Drew, J. A., Richards, S., Van Riper, D.,
& Williams, K. C. W. (2025). *IPUMS CPS: Version 13.0* [dataset].
IPUMS. <https://doi.org/10.18128/D030.V13.0>

The microdata are not redistributed, in accordance with the IPUMS license.
To replicate the analyses, register at <https://cps.ipums.org/>.


