---
title: "Tutorial 7"
subtitle: "Probit and Logit Models"
author: "Andrea Recine"
date: today
lang: en
---

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(modelsummary)
library(fixest)
library(wooldridge)
library(marginaleffects)
```



## Probit and Logit Models

In the previous tutorial, we studied the linear probability model (LPM) for models with a binary dependent variable. This tutorial introduces index models. In Probit and Logit models, the probability is a nonlinear function of the regressors that is constrained to lie between 0 and 1. This nonlinear function, which differs between Probit and Logit, depends on the linear index $X\beta$.

### Probit

**Probit model**: It uses the standard normal cumulative distribution function:
$$
   \Pr(Y=1|X_1,\dots, X_k) = \Phi(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + ... + \beta_k X_k) = \Phi(X\beta)
$$

where $\Phi(z)$ is the standard normal cumulative distribution function.

$$
\Phi(z) = \int_{-\infty}^{z} \frac{1}{\sqrt{2\pi}} e^{-\frac{t^2}{2}} dt
$$



The marginal effect of a variable $X_j$ in a Probit model is:

$$
\frac{\partial \Pr(Y=1|X_1,\dots, X_k)}{\partial X_j} = \phi(X\beta) \cdot \beta_j
$$

It is largest when $X\beta = 0$, that is, when the predicted probability is 0.5.


In both Logit and Probit models, the estimated coefficients **do not directly represent** marginal effects on the probability. The coefficient $\beta_j$ indicates the sign of the marginal effect, but not its magnitude. Unlike in the LPM, the magnitude depends on all explanatory variables and therefore varies across individuals.


We use the same regressors as in the LPM to estimate a Probit model. Recall that:

| Variable   | Description                                                                               |
| :---------- | :---------------------------------------------------------------------------------------- |
| $abuse$     | Dummy variable (1 = alcohol abuse, 0 = no alcohol abuse).                                  |
| $employ$    | Binary dependent variable (1 = employed, 0 = not employed).                              |
| $married$   | Dummy variable (1 = married, 0 = not married).                    |
| $educ$      | Years of education.       |
| $age$       | Age in years.       |

```{r}
data(alcohol, package = "wooldridge")
probit_alcohol <- feglm(employ ~ abuse + educ + age + married, data = alcohol, family = 'probit')
modelsummary(list("Employ (Probit)" = probit_alcohol), gof_omit = "AIC|BIC|RMSE")
```


The Probit coefficients represent effects on the latent index $X\beta$, measured in standard-normal standard deviation units.

## Logit Model


**Logit model**: It uses the logistic function to model the probability:
$$
   \Pr(Y=1|X_1,\dots, X_k) = \frac{e^{\beta_0 + \beta_1 X_1 + \beta_2 X_2 + ... + \beta_k X_k}}{1 + e^{\beta_0 + \beta_1 X_1 + \beta_2 X_2 + ... + \beta_k X_k}} = \Lambda(X\beta)
$$


where $\Lambda(z) = \frac{e^z}{1 + e^z}$ is the logistic function.


As in the Probit model, the marginal effect in the Logit model is largest when the predicted probability is 0.5, that is, when $X\beta = 0$.


Probit and Logit models often produce similar results. They may differ in the tails, however, because the logistic distribution has heavier tails than the normal distribution.

```{r}
logit_alcohol <- feglm(employ ~ abuse + educ + age + married, data = alcohol, family = 'logit')

modelsummary(list("Employ (Logit)" = logit_alcohol),
            gof_omit = "AIC|BIC|RMSE")

```



### Comparing the Curves

To illustrate the differences among the LPM, Probit, and Logit models, we plot the three functions $\Pr(Y=1 \mid X\beta)$ against the linear index $X\beta$:

```{r}
z <- seq(-5, 5, length.out = 400)
curve_df <- tibble(
  z = z,
  Probit = pnorm(z),
  Logit  = plogis(z),
  LPM    = 0.5 + 0.1 * z
) |>
  pivot_longer(-z, names_to = "Model", values_to = "p")

ggplot(curve_df, aes(x = z, y = p, color = Model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c("LPM" = "skyblue",
                                "Probit" = "darkblue",
                                "Logit" = "lightblue")) +
  labs(x = expression(X*beta), y = "Pr(Y = 1)",
       title = "LPM, Probit, and Logit") +
  theme_minimal()
```

The figure shows that:

- the **LPM** is linear, so for extreme values of $X\beta$ its predicted probability can fall outside $[0,1]$;
- **Probit** and **Logit** predictions always remain in $[0,1]$;
- the two curves are very close around $X\beta = 0$ (probability $\approx 0.5$), but the **Logit** has heavier tails and approaches 0 and 1 more slowly than the Probit.




### Predicted Probability

After estimating the Probit model, we can calculate the predicted probability for an individual in two steps:

1. construct the **linear index**
$$
X\beta = \beta_0 + \beta_{abuse}\cdot abuse + \beta_{educ}\cdot educ + \beta_{age}\cdot age + \beta_{married}\cdot married;
$$

2. apply the standard normal CDF:

$$
\Pr(employ = 1 \mid X) = \Phi(X\beta).
$$

::: {.callout-important title="Sign of $X\beta$ and $\Phi(X\beta) \gtrless 0.5$"}
The function $\Phi$ is the standard normal CDF. It is **strictly increasing**, and $\Phi(0) = 0.5$. Therefore:

- $X\beta > 0 \;\Longleftrightarrow\; \Phi(X\beta) > 0.5$;
- $X\beta < 0 \;\Longleftrightarrow\; \Phi(X\beta) < 0.5$.

The **sign** of $X\beta$ is sufficient to determine whether the predicted probability is above or below 0.5.

The same result holds for the Logit model, which uses the logistic function $\Lambda$ instead of $\Phi$. The logistic function is also strictly increasing and satisfies $\Lambda(0) = 0.5$.

:::

We extract the coefficients from the estimated Probit model:

```{r}
b <- coef(probit_alcohol)
b
```

Consider two individuals with contrasting profiles:

- **Individual A**: age $= 35$, education $= 14$ years, married, and **does not** abuse alcohol.
- **Individual B**: age $= 55$, education $= 4$ years, not married, and abuses alcohol.

**Individual A**:

```{r}
xb_A <- b["(Intercept)"] + b["abuse"]*0 + b["educ"]*14 +
        b["age"]*35 + b["married"]*1
unname(xb_A)

prob_A <- pnorm(xb_A)
unname(prob_A)
```

**Individual B**:

```{r}
xb_B <- b["(Intercept)"] + b["abuse"]*1 + b["educ"]*4 +
        b["age"]*55 + b["married"]*0
unname(xb_B)

prob_B <- pnorm(xb_B)
unname(prob_B)
```


We verify that the values for the two individuals match those obtained automatically with `predict()`:

```{r}
predict(probit_alcohol,
        newdata = data.frame(abuse   = c(0,  1),
                             educ    = c(14, 4),
                             age     = c(35, 55),
                             married = c(1,  0)))
```

We plot the two individuals on the $\Phi$ curve, highlighting the 0.5 threshold and the vertical line $X\beta = 0$:

```{r}
points <- tibble(
  individual = c("A", "B"),
  xb        = c(xb_A, xb_B),
  p         = c(prob_A, prob_B)
)

ggplot(tibble(z = seq(-3, 3, length.out = 400)),
       aes(x = z, y = pnorm(z))) +
  geom_line(color = "darkblue", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50") +
  geom_point(data = points, aes(x = xb, y = p),
             color = "skyblue", size = 3) +
  geom_text(data = points, aes(x = xb, y = p, label = individual),
            nudge_y = 0.05, fontface = "bold") +
  labs(x = expression(X * beta), y = expression(Phi(X * beta)),
       title = "Linear Index and Predicted Probability") +
  theme_minimal()
```

Individual A lies on the right-hand branch of the curve ($X\beta > 0$, $\Phi > 0.5$), whereas Individual B lies on the left-hand branch ($X\beta < 0$, $\Phi < 0.5$). The dashed lines $X\beta = 0$ and $\Phi = 0.5$ intersect at the symmetry point of $\Phi$, where the curve has its greatest slope.



### Logit: Odds

Logit coefficients represent effects on the log odds, not on the probability. Exponentiating a coefficient yields an **odds ratio**.


If $p(X_i) = \Lambda(X_i\beta)$, the odds are defined as:

$$
\text{odds}(X_i) = \frac{p(X_i)}{1 - p(X_i)} = e^{X_i\beta}
$$


Taking the logarithm of the odds gives:
$$
\log(\text{odds}(X_i)) = X_i\beta
$$

The log odds are a linear function of the explanatory variables, and each coefficient $\beta_j$ represents the effect of a one-unit increase in $X_j$ on the log odds that $Y=1$.


If $X_j$ increases by one unit while the other variables are held fixed, the ratio of the new odds to the initial odds is the **odds ratio**:

$$
\text{OR}_j = \frac{\text{odds}(X_j + 1)}{\text{odds}(X_j)} = \frac{e^{X\beta + \beta_j}}{e^{X\beta}} = e^{\beta_j}
$$

The odds ratio depends only on $\beta_j$, not on the initial value of $X_j$ or the other variables. It is therefore **constant across values of $X$**.


We calculate the odds ratios for the Logit model estimated using the `alcohol` data:

```{r}
exp(coef(logit_alcohol)) |> round(3)
```

**Interpretation**:

- The odds ratio for $abuse$ is 0.8. Alcohol abuse reduces the odds of being employed by approximately 20% (0.8 - 1 = -0.2) relative to no alcohol abuse.


- The odds ratio for $educ$ is 1.18. One additional year of education **increases** the odds of being employed by approximately 18% (1.18 - 1 = 0.18).




::: {.callout-important title="Odds ratio"}

An odds ratio is not a ratio of probabilities. An odds ratio of 2 does not mean that the probability doubles; it means that the ratio of successes to failures doubles. The Logit odds ratio is constant across values of $X_i$, whereas the marginal effect on the probability is not.
:::



## Average Marginal Effects (AME) and Marginal Effects at the Mean (MEM)



Marginal effects are essential for interpreting Logit and Probit models correctly.

**Average marginal effects (AME)**: Calculate the marginal effect for every observation and then average these individual effects over the sample.

Mathematically, the average marginal effect is:
$$
  \overline{\text{AME}} = \frac{1}{n} \sum_{i=1}^{n} \text{ME}_i
$$


where $\text{ME}_i$ is the marginal effect for observation $i$, and $n$ is the total number of observations.


**Marginal effects at the mean (MEM)**: Calculate marginal effects by evaluating all explanatory variables at their sample means.



::: {.callout-note title="Note"}
In the LPM, the marginal effect is simply the coefficient $\beta_j$:
$$
\frac{\partial \Pr(Y=1|X_1,\dots, X_k)}{\partial X_j} = \beta_j
$$
:::




### Manual Calculation


When $X_j$ is a dummy variable, its marginal effect is the difference between the predicted probabilities when $X_j=1$ and $X_j=0$. We calculate the marginal effect of `abuse` in the Probit and Logit models as follows.


Marginal effect of alcohol abuse in the Probit model:

```{r}
# Create two datasets: one with abuse=1 for everyone and one with abuse=0
alcohol_tmp1 <- alcohol
alcohol_tmp0 <- alcohol

# Set abuse=1 for everyone in the first dataset
alcohol_tmp1$abuse <- 1
# Set abuse=0 for everyone in the second dataset
alcohol_tmp0$abuse <- 0

# Calculate predicted probabilities for the Probit model
proba_employ_probit <- predict(probit_alcohol, newdata = alcohol_tmp1)
proba_nonemploy_probit <- predict(probit_alcohol, newdata = alcohol_tmp0)

# Calculate the average marginal effect (average predicted-probability difference)
ame_manual_probit <- mean(proba_employ_probit - proba_nonemploy_probit)
print(paste("Manual average marginal effect (Probit):", round(ame_manual_probit, 4)))
```


For the Logit model:

```{r}
# Calculate predicted probabilities for the Logit model
proba_employ_logit <- predict(logit_alcohol, newdata = alcohol_tmp1)
proba_nonemploy_logit <- predict(logit_alcohol, newdata = alcohol_tmp0)

# Calculate the average marginal effect (average predicted-probability difference)
ame_manual_logit <- mean(proba_employ_logit - proba_nonemploy_logit)
print(paste("Manual average marginal effect (Logit):", round(ame_manual_logit, 4)))
```


- According to the Probit model, alcohol abuse reduces the probability of being employed by 1.9 percentage points.
- According to the Logit model, alcohol abuse reduces the probability of being employed by 2.1 percentage points.

The average marginal effect accounts for how the effect of $abuse$ varies with individuals' other observed characteristics.


### Using `marginaleffects`

The average marginal effects calculated with `marginaleffects` equal those calculated manually above. The package also reports standard errors, which permit statistical inference. By default, these standard errors are calculated using the delta method.

The `avg_slopes` function without the `newdata` argument calculates the marginal effect for each observation and then averages the effects. This approach accounts for the empirical distribution of the explanatory variables in the sample.


```{r}
all_effects_logit <- avg_slopes(logit_alcohol)
all_effects_probit <- avg_slopes(probit_alcohol)
```



Average marginal effects (AME) for the Logit model:


```{r}
all_effects_logit
```

and for the Probit model:

```{r}
all_effects_probit
```


```{r}
# Table produced with modelsummary
modelsummary(list("Logit (AME)" = all_effects_logit,
            "Probit (AME)" = all_effects_probit), gof_omit = "AIC|BIC|RMSE|R2|Adj")
```


The effects are close to those estimated with the LPM.


### Marginal Effects at the Mean (MEM)

As an alternative to the AME, we can calculate the marginal effect by setting the explanatory variables to their **sample means**. With `marginaleffects`, pass `newdata = "mean"`:

```{r}
mem_probit <- avg_slopes(probit_alcohol, newdata = "mean")
mem_logit  <- avg_slopes(logit_alcohol,  newdata = "mean")

modelsummary(
  list("Probit (AME)" = all_effects_probit,
       "Probit (MEM)" = mem_probit,
       "Logit (AME)"  = all_effects_logit,
       "Logit (MEM)"  = mem_logit),
  gof_omit = "AIC|BIC|RMSE|R2|Adj"
)
```



To illustrate the difference, we calculate each variable's marginal effect **for every observation** in the sample (using `slopes()` instead of `avg_slopes()`) and superimpose the marginal effect at the mean (MEM):


```{r}
#| fig-height: 5
slopes_logit <- slopes(logit_alcohol)

ggplot(slopes_logit, aes(x = estimate)) +
  geom_histogram(bins = 60, fill = "lightblue", color = "white") +
  geom_vline(data = mem_logit, aes(xintercept = estimate),
             color = "darkblue", linewidth = 0.8) +
  facet_wrap(contrast ~ term, scales = "free") +
  labs(title = "Distribution of Individual Marginal Effects (Logit)",
       subtitle = "The vertical line is the marginal effect at the mean (MEM)",
       x = "Marginal effect",
       y = "Frequency") +
  theme_minimal()
```

Each histogram shows the distribution of individual marginal effects ($\text{ME}_i$) across sample observations. The **AME** is their average, whereas the **MEM** (vertical line) is the effect at a single point where all $X$ variables equal their means. The AME and MEM can differ substantially when the distribution is asymmetric or has long tails.


The same plot with the **AME** (the average of the individual effects) as the vertical line:

```{r}
#| fig-height: 5
ggplot(slopes_logit, aes(x = estimate)) +
  geom_histogram(bins = 60, fill = "lightblue", color = "white") +
  geom_vline(data = all_effects_logit, aes(xintercept = estimate),
             color = "darkblue", linewidth = 0.8) +
  facet_wrap(contrast ~ term, scales = "free") +
  labs(title = "Distribution of Individual Marginal Effects (Logit)",
       subtitle = "The vertical line is the average marginal effect (AME)",
       x = "Marginal effect",
       y = "Frequency") +
  theme_minimal()
```


The AME lies at the distribution's arithmetic mean, whereas the MEM corresponds to $\phi(X\beta)\cdot \beta_j$ evaluated at one point. In this dataset, the AME and MEM are numerically close for all variables. The MEM is **less interpretable for dummy variables** (`abuse`, `married`), because a dummy variable's mean is fractional and does not describe an actually observed individual.


**Interpretation** of the Probit average marginal effects (AME):


- Alcohol abuse reduces the probability of being employed by 1.9 percentage points.

- One additional year of education increases the probability of being employed by 1.5 percentage points.

- Being married increases the probability of being employed by 8.9 percentage points.





## Comparing the LPM, Probit, and Logit Models

This section compares the LPM estimates with the average marginal effects (AME) from Probit and Logit models for the `alcohol` and `crime1` datasets.

### The `alcohol` Dataset

```{r}
reg_lpm_alcohol <- feols(employ ~ abuse + educ + age + married, data = alcohol, vcov = "hetero")
modelsummary(list("Employ (LPM)" = reg_lpm_alcohol, "Employ AME (Probit)" = all_effects_probit, "Employ AME (Logit)" = all_effects_logit), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


**Interpretation of the results**:


The three models produce very similar results:

- Alcohol abuse is associated with an approximately 1.8-percentage-point lower probability of employment, holding age, years of education, and marital status fixed. The Probit and Logit average marginal effects indicate reductions of 1.9 and 2.1 percentage points, respectively.




- One additional year of education is associated with a 1.7-percentage-point increase in the probability of employment, holding the other variables fixed. The Probit and Logit average marginal effects indicate increases of 1.5 and 1.4 percentage points, respectively.




### The `crime1` Dataset



The variables are:


| Variable   | Description                                                                               |
| :---------- | :---------------------------------------------------------------------------------------- |
| $pcnv$     | Proportion of previous arrests that resulted in conviction.                      |
| $avgsen$    | Average sentence served for previous convictions (months).                              |
| $tottime$   | Total months spent in prison since age 18 and before 1986.                    |
| $ptime86$     | Months spent in prison in 1986.       |
| $qemp86$     | Number of quarters in which the individual was employed in 1986.       |
| $black$     | Dummy variable (1 = Black, 0 = otherwise).       |
| $hispan$     | Dummy variable (1 = Hispanic, 0 = otherwise).       |



```{r}
data("crime1", package = "wooldridge")
head(crime1)


crime1 <- crime1 %>%
  mutate(arrested = ifelse(narr86 > 0, 1, 0))



reg_lpm_crime <- feols(arrested ~ pcnv + avgsen + tottime + ptime86 + qemp86 + black + hispan, data = crime1, vcov = "hetero")
modelsummary(list("Arrested" = reg_lpm_crime), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```


The Probit and Logit models for the `crime1` dataset are:

```{r}
probit_crime <- feglm(arrested ~ pcnv + avgsen + tottime + ptime86 + qemp86 + black + hispan, data = crime1, family = 'probit')
logit_crime <- feglm(arrested ~ pcnv + avgsen + tottime + ptime86 + qemp86 + black + hispan, data = crime1, family = 'logit')

modelsummary(list("Arrested (LPM)" = reg_lpm_crime, "Arrested (Probit)" = probit_crime, "Arrested (Logit)" = logit_crime), gof_omit = "AIC|BIC|RMSE|R2 Adj.")

```

We cannot interpret the Probit and Logit coefficients directly, but we can calculate average marginal effects (AME) and compare them with the LPM coefficients:


```{r}
ame_crime_probit <- avg_slopes(probit_crime)
ame_crime_logit <- avg_slopes(logit_crime)

ame_crime_probit
ame_crime_logit
```




```{r}
modelsummary(list("Arrested (LPM)" = reg_lpm_crime, "Arrested (Probit AME)" = ame_crime_probit, "Arrested (Logit AME)" = ame_crime_logit), gof_omit = "AIC|BIC|RMSE|R2 Adj.")
```

**Interpretation of the results**:



- A 0.01 increase in $pcnv$ (a one-percentage-point increase in the share of previous arrests resulting in conviction) is associated with a 0.152-percentage-point decrease in the probability of arrest, holding the other variables fixed. The Probit and Logit average marginal effects imply decreases of 0.165 and 0.164 percentage points, respectively.


- One additional quarter of employment in 1986 is associated with a 3.8-percentage-point decrease in the probability of arrest, holding the other variables fixed. The Probit and Logit average marginal effects imply decreases of 3.8 and 3.7 percentage points, respectively.



- Holding the other variables fixed, a Black individual has a 17.0-percentage-point higher probability of arrest than an otherwise comparable non-Black individual. The Probit and Logit average marginal effects for $black$ are virtually identical to the LPM coefficient (17.0 percentage points in both models).
